---
title: "Relative_abundance"
output: html_document
---
This script aims to analyse the relative abundance of Pocillopora around the island of Mo’orea under 12 environmental conditions. This script also aims to characterize the benthos at each site in order to better understand the biodiversity in which the Pocillopora community lives.

Q1: What does the benthic cover of the lagoon around the island look like?
Q2: How are Pocillopora species distributed around the lagoon?

What corals are found on the lagoon reefs of Mo’orea?
What Pocillopora species are found on the lagoon reefs of Mo’orea?

H0: Benthic cover, coral cover, and Pocillopora cover differ by site.

#Install and load packages
```{r}
library("ggplot2")
library("ggpubr")
library("tidyverse")
```
```{r}
data_haplotype <- read.csv("/Users/pierrickharnay/Documents/MyProjects/Pocillopora_under_environmental_stressors-/Data/Haplotype_data_modified_duplication.csv")

data_size <- read.csv("/Users/pierrickharnay/Documents/MyProjects/Pocillopora_under_environmental_stressors-/Data/POC_RA_Size.csv")

data_quadrats <- read.csv("/Users/pierrickharnay/Documents/MyProjects/Pocillopora_under_environmental_stressors-/Data/20260226_Quadrat_Benthos_RA.csv")

data_haplotype
data_size
data_quadrats
```
```{r}
### DATA CLEANING ###

#formatting the dataset as we need 
#removing useless column 

#data Haplotype.
# Remove uninformative columns (positions 1, 2, 3, 4, 7, 9)
data_species <- data_haplotype |>
  select(-c(1, 2, 3, 4, 7, 9)) |>
  drop_na()

#data size.
# Remove uninformative columns (positions 4, 5, 6, 8)
data_diameters <- data_size |>
  select(-c(4, 5, 6, 8)) |>
  drop_na()


# Merge species and diameter datasets by shared Sample.ID
data <- data_species |>
  mutate(Sample.ID = as.integer(Sample.ID)) |>
  inner_join(data_diameters, by = "Sample.ID")

# Inspect problematic Sample.ID values in data_species
data_species |>
  filter(is.na(as.integer(Sample.ID))) |>
  distinct(Sample.ID)
#the issues come from 2 samples because .bis present at the end of the sample.ID

# Check for duplicate Sample.IDs in each dataset
data_species |> count(Sample.ID) |> filter(n > 1)
data_diameters |> count(Sample.ID) |> filter(n > 1)

# --- QC Check: Identify remaining duplicate Sample.IDs ------
dup_species   <- data_species   |> count(Sample.ID) |> filter(n > 1) |> pull(Sample.ID)
dup_diameters <- data_diameters |> count(Sample.ID) |> filter(n > 1) |> pull(Sample.ID)

# --- Remove remaining duplicates ----------------------------
data_species   <- data_species   |> filter(!Sample.ID %in% dup_species)
data_diameters <- data_diameters |> filter(!Sample.ID %in% dup_diameters)

```

```{r}
# ============================================================
# QUALITY CONTROL
# ============================================================
# This section verifies the integrity of the datasets before
# any analysis. Each check is documented for reproducibility
# and transparency for reviewers.

# --- QC 1: Dataset Dimensions -------------------------------
# Verify the number of rows and columns after cleaning
cat("data_species dimensions:  ", dim(data_species),  "\n")
cat("data_diameters dimensions:", dim(data_diameters), "\n")

# --- QC 2: Dataset Structure --------------------------------
# Verify column names and types are as expected
str(data_species)
str(data_diameters)

# --- QC 3: Missing Values -----------------------------------
# Confirm no NAs remain after drop_na() in cleaning step
cat("\nMissing values in data_species:\n")
colSums(is.na(data_species))

cat("\nMissing values in data_diameters:\n")
colSums(is.na(data_diameters))

# --- QC 4: Duplicate Sample.IDs -----------------------------
# Confirm no duplicates remain after deduplication step
cat("\nRemaining duplicates in data_species:\n")
data_species  |> count(Sample.ID) |> filter(n > 1)

cat("\nRemaining duplicates in data_diameters:\n")
data_diameters |> count(Sample.ID) |> filter(n > 1)

# --- QC 5: Value Range & Consistency ------------------------
# Check for impossible or unexpected values in key variables
cat("\nSummary of data_diameters:\n")
summary(data_diameters)

cat("\nUnique species found in data_species:\n")
unique(data_species$Species)

cat("\nUnique haplotypes found in data_species:\n")
unique(data_species$Haplotype)

# --- QC 6: Join Integrity -----------------------------------
# Identify Sample.IDs present in one dataset but not the other
# These samples will be excluded from the merged dataset
lost_from_species <- anti_join(
  data_species |> mutate(Sample.ID = as.integer(Sample.ID)),
  data_diameters,
  by = "Sample.ID"
)

lost_from_diameters <- anti_join(
  data_diameters,
  data_species |> mutate(Sample.ID = as.integer(Sample.ID)),
  by = "Sample.ID"
)

# Store excluded Sample.IDs for traceability
excluded_from_species   <- lost_from_species  |> pull(Sample.ID)
excluded_from_diameters <- lost_from_diameters |> pull(Sample.ID)

cat("\nSamples excluded from data_species (no size match):",   length(excluded_from_species),   "\n")
cat("Samples excluded from data_diameters (no species match):", length(excluded_from_diameters), "\n")

# --- Remove unmatched samples from each dataset -------------
data_species <- data_species |>
  filter(!Sample.ID %in% excluded_from_species)

data_diameters <- data_diameters |>
  filter(!Sample.ID %in% as.character(excluded_from_diameters))

# --- Rebuild merged dataset after exclusions ----------------
data <- data_species |>
  mutate(Sample.ID = as.integer(Sample.ID)) |>
  inner_join(data_diameters, by = "Sample.ID")

cat("\nFinal merged dataset dimensions after exclusions:", dim(data), "\n")

# --- Remove .bis samples & rebuild clean datasets -----------
data_species <- data_species |>
  filter(!grepl("\\.bis$", Sample.ID, ignore.case = TRUE)) |>
  mutate(Sample.ID = as.integer(Sample.ID)) |>
  filter(!is.na(Sample.ID))                                   # safety net for any remaining coercion issues

# --- Rebuild merged dataset after exclusions ----------------
data <- data_species |>
  inner_join(data_diameters, by = "Sample.ID")

# --- Confirm no coercion warnings remain --------------------
cat("\nFinal merged dataset dimensions after exclusions:", dim(data), "\n")
cat("Missing values in final merged dataset:\n")
colSums(is.na(data))

# Check all unique species names
sort(unique(data$Species))
sort(unique(data$Haplotype))

sort(unique(data$Site))
sort(unique(data$Transect))

# Cross-check: how many transects per site?
data |>
  group_by(Site) |>
  summarise(n_transects = n_distinct(Transect),
            n_samples   = n())

#check diameters values 
summary(data$Diameter.mean)

# Flag potential outliers (e.g. > 500cm is biologically impossible)
data |> filter(Diameter.mean > 500 | Diameter.mean <= 0)

#bcs is stored as character instead of numeric 
# Convert Diameter.mean to numeric
data <- data |>
  mutate(Diameter.mean = as.numeric(Diameter.mean))

# Verify the conversion
summary(data$Diameter.mean)

# Flag potential outliers (biologically impossible values)
data |> filter(Diameter.mean <= 0 | Diameter.mean > 500)

# Confirm no NAs introduced by the numeric conversion
cat("NAs in Diameter.mean after conversion:", sum(is.na(data$Diameter.mean)), "\n")

# Confirm the range is biologically reasonable for Pocillopora
summary(data$Diameter.mean)

# Flag potential outliers
data |> filter(Diameter.mean <= 0 | Diameter.mean > 500)

# Visual check of the diameter distribution
hist(data$Diameter.mean, 
     breaks = 50,
     main   = "Distribution of Diameter.mean",
     xlab   = "Diameter (cm)",
     col    = "steelblue")

# --- QC: Handle colonies too large to measure ---------------
# Colonies recorded as 100cm represent colonies whose diameter
# could not be measured in the field (> 100cm). These are
# reclassified as a distinct ">100cm" size class rather than
# being removed, as they represent real biological observations.

data <- data |>
  mutate(LifeStage = case_when(
    Diameter.mean == 100 ~ ">100cm",
    Diameter.mean  <   2 ~ "0-2cm",
    Diameter.mean  <   5 ~ ">2cm-5cm",
    Diameter.mean  <  14 ~ ">5cm-14cm",
    Diameter.mean  <  100 ~ ">14cm",
    .default = NA_character_
  )) |>
  mutate(LifeStage = factor(LifeStage, 
                            levels = c("0-2cm", ">2cm-5cm", ">5cm-14cm", 
                                       ">14cm", ">100cm")))

# Verify the distribution across size classes
data |>
  group_by(LifeStage) |>
  summarise(n = n(), .groups = "drop")

#because some NA have been find into the dataset, we will remove them to avoid any trouble during the stats analysis 
# Remove rows with NA in LifeStage
data <- data |>
  filter(!is.na(LifeStage))

# Confirm no NAs remain
cat("Remaining NAs in LifeStage:", sum(is.na(data$LifeStage)), "\n")
cat("Final dataset dimensions:", dim(data), "\n")
```











