---
title: "pocillopora_abundance"
subtitle: "Statistical analysis and visualization pipeline"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    embed-resources: true
    fig-width: 10
    fig-height: 7
execute:
  warning: false
  message: false
editor: visual
---

This script contains the data cleaning and plotting of benthic cover and Pocillopora around the island of Mo’orea at 11 sites. 

#load packages

```{r}
#| label: setup
#| include: false
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r}
library("ggplot2")
library("vegan")
library("tidyverse")
library("purrr")
library("ggsn")
library("colorspace")
library("patchwork")
library("ggmap")
library("ggh4x")
```

Load Data Files for Analysis

```{r}
#read in species id data
data_species <- read.csv("Data/species_id.csv")

#read in colony diameter data
data_size <- read.csv("Data/colony_size.csv")

#read in benthic quadrat data
data_benthic <- read.csv("Data/quadrat_benthos.csv")
```

Data cleaning of species ID and colony diameter

```{r}
#confirm the data re filtered to only data from the first 3 transects
data_size <- data_size %>%
  filter(Transect %in% c("T1", "T2", "T3"))

#check data_size to see if there are any values that are 100 in all Diameter columns exactly that do not have text in the Note column. The 100 indicates >14cm because the whole colony was not in the photograph and it was longer than the size standards of 15 marked cm
data_size |>
  dplyr::filter(Diameter.1 == 100 & Diameter.2 == 100 & Diameter.3 == 100 & (is.na(Note) | Note == ""))

# Remove uninformative columns ("Diameter.1","Diameter.2","Diameter.3","Note") and filter out NA rows and DIV/0! rows and check if diameter is numeric
data_diameters <- data_size %>%
  mutate(Diameter.mean = rowMeans(across(c(Diameter.1, Diameter.2, Diameter.3)), na.rm = TRUE)) %>%
  select(-c("Diameter.1","Diameter.2","Diameter.3","Note")) %>%
  na.omit() %>%
  filter(Diameter.mean!="#DIV/0!") 

str(data_diameters)

```

Quality Control of Sample.IDs

```{r}
# Check species and diameter datasets by Sample.ID to see if there are any ID problems
# Remove uninformative columns ("Sanger.Plate.Name","Sequence.ID","Plate.Well", "gDNA.Extraction.Plate") and filter out NA rows
data_species <- data_species %>%
  select(-c("Sanger.Plate.Name","Sequence.ID","Plate.Well", "gDNA.Extraction.Plate", "Notes")) %>% na.omit()

# Check to see IDs are all distinct
dup_species <- data_species %>% count(Sample.ID, sort = TRUE) %>% filter(n > 1)
dup_species$Sample.ID <- as.numeric(dup_species$Sample.ID)
  
#generate a list of duplicated IDs
dup_rows_species <- data_species %>%
  filter(Sample.ID %in% dup_species$Sample.ID) %>%
  arrange(Sample.ID)

#save a file of problematic Sample.IDs in the species file to QC
write.csv(dup_rows_species, "Output/dup_rows_species.csv")

# Check to see they are all distinct
dup_diameters <-data_diameters %>% count(Sample.ID, sort = TRUE) %>% filter(n > 1)
dup_diameters$Sample.ID <- as.numeric(dup_diameters$Sample.ID)

#generate a list of duplicated IDs
dup_rows_size <- data_diameters %>%
  filter(Sample.ID %in% dup_diameters$Sample.ID) %>%
  arrange(Sample.ID)

#save a file of problematic Sample.IDs in the size file to QC
write.csv(dup_rows_size, "Output/dup_rows_size.csv")

# Check to see if all Sample.IDs are all numeric
bad_ids_species <- data_species$Sample.ID[
  is.na(suppressWarnings(
    as.numeric(as.character(data_species$Sample.ID))
  ))
]
bad_ids_species

#An issues come from any sample containing _BIS  at the end of the Sample.ID
#For these samples "_BIS" means #.2 in French, essentially a duplicate name accidentally given to extra samples
#We do not have pictures of _BIS corals so they are going to be excluded from the dataset

data_species <- data_species %>%
  filter(!grepl("^[0-9]+_BIS$", Sample.ID)) %>% #filter out _BIS samples
  mutate(Sample.ID = as.numeric(as.character(Sample.ID))) # change the sample.ID to numeric
  
# Check to see if all Sample.ID are all numeric
bad_ids_species <- data_species$Sample.ID[
  is.na(suppressWarnings(
    as.numeric(as.character(data_species$Sample.ID))
  ))
]
bad_ids_species
#There are only numeric IDs in species data

# Check to see if all Sample.ID are all numeric
bad_ids_diameters <- data_diameters$Sample.ID[
  is.na(suppressWarnings(
    as.numeric(as.character(data_diameters$Sample.ID))
  ))
]
bad_ids_diameters 
#There are only numeric IDs in diameter data

# Remove remaining duplicates if QC and manual check of the photos and notebooks do not reconcile and we cannot confirm the sample ID
data_species <- data_species %>%
  filter(!Sample.ID %in% dup_species$Sample.ID)

data_diameters <- data_diameters %>% 
  filter(!Sample.ID %in% dup_diameters$Sample.ID)

```

Sanity checks for data quality

```{r}
# This section verifies the integrity of the datasets before
# any analysis. Each check is documented for reproducibility
# and transparency.

#Check Dataset Dimensions
# Verify the number of rows and columns after cleaning
cat("data_species dimensions:  ", dim(data_species),  "\n")
cat("data_diameters dimensions:", dim(data_diameters), "\n")
#when all species IDs are complete we would expect the same number of rows in both dataframes
#We have ~300 more species than we do sizes, which is likely due to the species coming from transects 4,5,6 and we are not able to filter them by transect becasue the file did not have metadata within it


#Check Dataset Structure
# Verify column names and types are as expected
#Species should be character
#Sample.ID should be numeric
#Diameter.mean should be numeric
str(data_species)
str(data_diameters)


#Check Missing Values
# Confirm no NAs remain after na.omit in cleaning step
#results from this should be all 0's
cat("\nMissing values in data_species:\n")
colSums(is.na(data_species))

cat("\nMissing values in data_diameters:\n")
colSums(is.na(data_diameters))


#Check Duplicate Sample.IDs
# Confirm no duplicates remain after deduplication step
cat("\nRemaining duplicates in data_species:\n")
data_species  %>% count(Sample.ID) %>% filter(n > 1)

cat("\nRemaining duplicates in data_diameters:\n")
data_diameters %>% count(Sample.ID) %>% filter(n > 1)

#Check Value Range and if Values are Realistic
# Check for impossible or unexpected values in key variables
#this should be numeric, not character
cat("\nSummary of data_diameters:\n")
range(data_diameters$Diameter.mean)
#this should range from >0-100, where 100 indicates >14cm because the whole colony was not in the photograph and it was longer than the size standards of 15 marked cm

cat("\nUnique sites found in data_diameters:\n")
unique(data_diameters$Site)

cat("\nUnique transects found in data_diameters:\n")
unique(data_diameters$Transect)

cat("\nUnique species found in data_species:\n")
#should be "P. grandis" "P. meandrina" "P. verrucosa" "P. effusa" "P. tuahiniensis" "P. acuta"  
unique(data_species$Species)

cat("\nUnique haplotypes found in data_species:\n")
#should be "Haplotype 1a" "Haplotype 3"  "Haplotype 2"  "Haplotype 11" "Haplotype 10" "Haplotype 8a" "Haplotype 5a"
unique(data_species$Haplotype)


```

Join the species and size datasets together for Size Class specific analyses

```{r}
##Check Data Join Integrity 
# Identify Sample.IDs present in one dataset but not the other
# These samples will be excluded from the merged dataset
lost_from_species <- anti_join(data_species,
  data_diameters,
  by = "Sample.ID"
)

lost_from_diameters <- anti_join(
  data_diameters,
  data_species,
  by = "Sample.ID"
)

# Store excluded Sample.IDs for QC as needed
excluded_from_species   <- lost_from_species  %>% pull(Sample.ID)
excluded_from_diameters <- lost_from_diameters %>% pull(Sample.ID)

cat("\nSamples excluded from data_species (no size match):",   length(excluded_from_species),   "\n")
cat("Samples excluded from data_diameters (no species match):", length(excluded_from_diameters), "\n")

# Remove unmatched samples from each dataset
data_species <- data_species %>%
  filter(!Sample.ID %in% excluded_from_species)
dim(data_species)

data_diameters <- data_diameters %>%
  filter(!Sample.ID %in% excluded_from_diameters)
dim(data_diameters)

#Rebuild merged dataset after any exclusions
data <- data_species %>%
  inner_join(data_diameters, by = "Sample.ID")

cat("\nFinal merged dataset dimensions after exclusions:", dim(data), "\n")

# Confirm there are no coercion warnings remaining
cat("\nFinal merged dataset dimensions after exclusions:", dim(data), "\n")
cat("Missing values in final merged dataset:\n")
colSums(is.na(data))

# Check all unique species names
sort(unique(data$Species))
sort(unique(data$Haplotype))

sort(unique(data$Site))
sort(unique(data$Transect))

# Cross-check: how many transects per site?
data  %>%
  group_by(Site)  %>%
  summarise(n_transects = n_distinct(Transect),
            n_samples   = n())


#Assign colonies to Size Classes
# Colonies recorded as 100cm represent colonies whose diameter
# could not be measured in the field. These are
# reclassified as a distinct 100 value rather than
# being removed, as they represent real biological observations. We only analyze these as a >14cm size class

data <- data %>%
  mutate(
    SizeClass = case_when(
      Diameter.mean <= 2 ~ "0-2cm",
      Diameter.mean > 2 & Diameter.mean <= 5 ~ ">2cm-5cm",
      Diameter.mean > 5 & Diameter.mean <= 14 ~ ">5cm-14cm",
      Diameter.mean > 14 ~ ">14cm",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    SizeClass = factor(
      SizeClass,
      levels = c("0-2cm", ">2cm-5cm", ">5cm-14cm", ">14cm")
    )
  )

# Remove rows with NA in SizeClass
data <- data %>%
  filter(!is.na(SizeClass))

# Confirm no NAs remain
cat("Remaining NAs in SizeClass:", sum(is.na(data$SizeClass)), "\n")
cat("Remaining NAs in Site:", sum(is.na(data$Site)), "\n")
cat("Remaining NAs in Transect:", sum(is.na(data$Transect)), "\n")
cat("Final dataset dimensions:", dim(data), "\n")
```

Quantify Pocillopora species density

```{r}
#for the rest of the analysis we add the shore for species distribution analysis
# Manually assign each site to its corresponding shore
# based on the geographic position around Mo'orea island
# North: sites S2, S3, S4, S5
# East:  sites S6, S9, S11
# West:  sites S12, S14, S15, S18

#Add shore category based on site information
data <- data %>%
  mutate(Shore = case_when(
    Site %in% c("S2", "S3", "S4", "S5")       ~ "North",
    Site %in% c("S6", "S9", "S11")         ~ "East",
    Site %in% c("S12", "S14", "S15", "S18")   ~ "West",
    .default = NA_character_
  )) %>%
  mutate(Shore = factor(Shore, levels = c("North", "East", "West")))

# Verify shore assignment
data %>%
  group_by(Shore, Site) %>%
  summarise(n_samples = n(), .groups = "drop") %>%
  arrange(Shore, Site)

# Confirm no unassigned sites
cat("Samples with no shore assigned:", sum(is.na(data$Shore)), "\n")

str(data)

#Calculate Species × SizeClass density by site
site_species_size_density <- data %>%
  count(
    Site,
    Species,
    SizeClass,
    name = "N_colonies"
  ) %>%
  mutate(
    Density_m2 = N_colonies / 75
  )

site_species_size_density

site_species_size_density <- site_species_size_density %>%
  rename(site.name = Site)

write.csv(site_species_size_density, "Output/pocillopora_site_species_size_density.csv")


# Calculate density of colonies per m2 so we can quantify mean and sem
transect_area <- 25

density_summary <- data %>%
  count(
    Site,
    Transect,
    Species,
    SizeClass,
    name = "N"
  ) %>%
  complete(
    Site,
    Transect,
    Species,
    SizeClass,
    fill = list(N = 0)
  ) %>%
  mutate(
    Density_m2 = N / transect_area
  ) %>%
  group_by(
    Site,
    Species,
    SizeClass
  ) %>%
  summarise(
    MeanDensity = mean(Density_m2),
    SDDensity   = sd(Density_m2),
    SE          = sd(Density_m2) / sqrt(n()),
    N_transects = n(),
    .groups = "drop"
  )

# Set factor orders
density_summary$Species <- factor(
  density_summary$Species,
  levels = c(
    "P. verrucosa",
    "P. tuahiniensis",
    "P. meandrina",
    "P. grandis",
    "P. effusa",
    "P. acuta"
  )
)

density_summary$SizeClass <- factor(
  density_summary$SizeClass,
  levels = c(
    "0-2cm",
    ">2cm-5cm",
    ">5cm-14cm",
    ">14cm"
  )
)

density_summary$Site <- factor(
  density_summary$Site,
  levels = c(
    "S2", "S3", "S4", "S5",
    "S6", "S9", "S11",
    "S12", "S14", "S15", "S18"
  )
)

# Facet labels (italic species names)
species_labs <- c(
  "P. acuta"        = "italic('P. acuta')",
  "P. effusa"       = "italic('P. effusa')",
  "P. grandis"      = "italic('P. grandis')",
  "P. meandrina"    = "italic('P. meandrina')",
  "P. tuahiniensis" = "italic('P. tuahiniensis')",
  "P. verrucosa"    = "italic('P. verrucosa')"
)

# Colors
species_cols <- c(
  "P. tuahiniensis" = "#D55E00",
  "P. meandrina"    = "#0072B2",
  "P. verrucosa"    = "#E69F00",
  "P. grandis"      = "#56B4E9",
  "P. effusa"       = "#009E73",
  "P. acuta"        = "#e63946"
)

# Legend labels (italic species names)
species_labels <- setNames(
  lapply(
    names(species_cols),
    function(x) bquote(italic(.(x)))
  ),
  names(species_cols)
)

# Plot
pocillopora.density.plot <- ggplot(
  density_summary,
  aes(
    x = Site,
    y = MeanDensity,
    fill = Species
  )
) +
  geom_col() +
  geom_errorbar(
    aes(
      ymin = MeanDensity - SE,
      ymax = MeanDensity + SE
    ),
    width = 0.2
  ) +
  facet_grid(
    Species ~ SizeClass,
    scales = "free",
    labeller = labeller(
      Species = as_labeller(
        species_labs,
        label_parsed
      )
    )
  ) +
  scale_fill_manual(
    values = species_cols,
    labels = species_labels
  ) +
  labs(
    x = "Site",
    y = expression("Density (colonies m"^{-2} * ")"),
    fill = "Species"
  ) +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 1
    ),
    strip.background = element_rect(
      fill = "gray",
      color = "black"
    ),
    strip.text.x = element_text(
      color = "black",
      face = "bold"
    ),
    strip.text.y = element_text(
      color = "black",
      face = "italic"
    ),
    legend.position = "none"
  )

pocillopora.density.plot

ggsave(
  filename = "Output/pocillopora.density.plot.jpg",
  plot = pocillopora.density.plot,
  device = "jpg",
  width = 8,
  height = 9,
  units = "in"
)

#write out density data
write.csv(density_summary,"Output/pocillopora.density.csv")

```

Plot the Moorea site map

```{r}
register_google(
  key = Sys.getenv("GOOGLE_MAPS_API_KEY")
)
# location
Moorea<-data.frame(lon = -149.83246425684064, lat = -17.531092816791094)

#Map base
M1<-get_map(Moorea,zoom = 12, maptype = 'satellite')

bbx <- c(left=-149.802,bottom= -17.480,right=-149.805,top=-17.475)
x <- c(bbx["left"], bbx["left"], bbx["right"], bbx["right"])
y <- c(bbx["bottom"], bbx["top"], bbx["top"], bbx["bottom"])
df <- data.frame(x, y)

sitedata <- read.csv("Data/site_coords.csv")
str(sitedata)

labels <- sitedata$Site

#plot the map
map_plot <- ggmap(M1) +
  geom_point(
    data = sitedata,
    aes(x = long, y = lat),
    size = 2,
    color = "yellow"
  ) +
  geom_text(
    data = sitedata,
    aes(x = long, y = lat, label = Site),
    vjust = -1,
    size = 3,
    color = "yellow"
  ) +
  scalebar(
    data = sitedata,
    dist = 2,
    dist_unit = "km",
    transform = TRUE,
    model = "WGS84",
    st.size = 2,
    st.color = "white",
    box.color = "white",
    height = 0.001,
    location = "bottomleft"
  ) +
  xlab("") +
  ylab("")

map_plot
```

Data Cleaning and QC for the benthic quadrat data
```{r}

data_benthic <- data_benthic %>%
  rename(Benthic.Type = Label.code)

#the naming structure should be Quadrat_Site_Transect.jpg or jpeg
#check to see if there are more than two _ in a Name
data_benthic %>%
  mutate(
    n_underscore = str_count(Name, "_"),
    prefix = str_extract(Name, "^[^_]+_[^_]+_[^_]+")
  ) %>%
  group_by(prefix) %>%
  mutate(prefix_count = n()) %>%
  ungroup() %>%
  filter(n_underscore >= 3) %>%
  select(Name, prefix, prefix_count) %>%
  distinct()
#There are more than two _ in a name for several, but they each have 100 points, which is the number of points counted in CoralNet

#remove the photo file extensions
data_benthic <- data_benthic %>%
  mutate(Name = sub("\\.JPG$", "", Name, ignore.case = TRUE)) %>%
  mutate(Name = sub("\\.jpeg$", "", Name, ignore.case = TRUE)) %>%
  separate(
    Name,
    into = c("Quadrat", "Site", "Transect"),
    sep = "_",
    remove = TRUE
  )

#filter to only T1-T3
data_benthic <- data_benthic %>%
  filter(Transect %in% c("T1", "T2", "T3")) %>% filter(!Site =="S1")

#check to see if all quadrat, site, and transect numbers exist
unique(data_benthic$Quadrat) 
unique(data_benthic$Site) 
unique(data_benthic$Transect) 

#set site order
site_order <- c(
  "S2","S3","S4","S5","S6","S9","S11","S12","S14","S15","S18"
)

#Add Shore information
data_benthic <- data_benthic %>%
  mutate(Shore = case_when(
    Site %in% c("S2", "S3", "S4", "S5")       ~ "North",
    Site %in% c("S6", "S9", "S11")         ~ "East",
    Site %in% c("S12", "S14", "S15", "S18")   ~ "West",
    .default = NA_character_
  )) %>%
  mutate(Shore = factor(Shore, levels = c("North", "East", "West")))

unique(data_benthic$Shore)

#count the number of Benthic.Type points by quadrat transect and site
data_benthic_counts <- data_benthic %>%
  count(Site, Transect, Quadrat, name = "Count")
#they should all be 100

#count the number of quadrats per transect and site
quadrat_counts <- data_benthic %>%
  distinct(Site, Transect, Quadrat) %>%
  count(Site, Transect, name = "N_Quadrats")
#they should be a maximum of 26

# Add ecological group for color mapping and filtering
data_benthic <- data_benthic %>%
  mutate(
    Benthic.Group = case_when(
      Benthic.Type %in% c(
        "Acanthastr", "Acropora", "Astreopora", "Cyphastrea",
        "Fungia", "Gardinoser", "Leptastrea", "Leptoseris",
        "Lobophylli", "Montastrea", "Montipora", "Pachyseris",
        "Pavona", "Pocillopo", "Porites", "Psammomora",
        "Tubastrea", "Sandolitha", "Un coral"
      ) ~ "Hard_Coral",

      Benthic.Type %in% c(
        "Macroalgae"
      ) ~ "Algae",
      
      Benthic.Type %in% c(
        "CCA_bare"
      ) ~ "CTB",

      Benthic.Type %in% c(
        "Sand", "Unk", "GiCl", "Millepora"
      ) ~ "Misc",

      TRUE ~ "Other"
    )
  )

# Map abbreviated CoralNet codes to full plotting names
benthic_name_map <- c(
  "Acanthastr" = "Acanthastrea",
  "Acropora"   = "Acropora",
  "Astreopora" = "Astreopora",
  "Cyphastrea" = "Cyphastrea",
  "Fungia"     = "Fungia",
  "Gardinoser" = "Gardineroseris",
  "Leptastrea" = "Leptastrea",
  "Leptoseris" = "Leptoseris",
  "Lobophylli" = "Lobophyllia",
  "Montastrea" = "Montastraea",
  "Montipora"  = "Montipora",
  "Pachyseris" = "Pachyseris",
  "Pavona"     = "Pavona",
  "Pocillopo"  = "Pocillopora",
  "Porites"    = "Porites",
  "Psammomora" = "Psammocora",
  "Tubastrea"  = "Tubastraea",
  "Sandolitha" = "Sandalolitha",
  "Un coral"   = "Unidentified coral",
  "Macroalgae" = "Macroalgae",
  "CCA_bare"   = "Rubble, Crustose Coralline Algae, Turf",
  "Sand"       = "Sand",
  "Unk"        = "Unknown",
  "GiCl"       = "Giant clam",
  "Millepora"  = "Millepora"
)

data_benthic <- data_benthic %>%
  mutate(
    Benthic.Name = recode(Benthic.Type, !!!benthic_name_map)
  )

# Calculate relative abundance
benthic_rel <- data_benthic %>%
  count(Benthic.Group, Benthic.Name, name = "Count") %>%
  mutate(RelAbundance = Count / sum(Count)) %>%
  arrange(Benthic.Group, desc(RelAbundance)) %>%
  mutate(
    Benthic.Name = factor(Benthic.Name, levels = unique(Benthic.Name))
  )

# Italicize genus names, keep non-genus labels plain
non_genus_names <- c(
  "Unidentified coral",
  "Macroalgae",
  "Rubble, Crustose Coralline Algae, Turf",
  "Sand",
  "Unknown",
  "Giant clam"
)

benthic_labels <- setNames(
  lapply(levels(benthic_rel$Benthic.Name), function(x) {
    if (x %in% non_genus_names) {
      x
    } else {
      bquote(italic(.(x)))
    }
  }),
  levels(benthic_rel$Benthic.Name)
)

# Calculate the Number of unique quadrats
N_quadrats <- data_benthic %>%
  distinct(Site, Transect, Quadrat) %>%
  nrow()

# Assign base colors by ecological group
group_bases <- c(
  Hard_Coral = "#332288",
  Algae      = "#117733",
  Misc       = "#888888",
  CTB      = "#DDCC77"
)

# Generate shades within each ecological group
benthic_cols <- benthic_rel %>%
  distinct(Benthic.Group, Benthic.Name) %>%
  group_by(Benthic.Group) %>%
  mutate(
    Color = lighten(
      group_bases[Benthic.Group[1]],
      amount = seq(0, 0.65, length.out = n())
    )
  ) %>%
  ungroup()

# map colors to groups
color_mapping <- setNames(
  benthic_cols$Color,
  benthic_cols$Benthic.Name
)

# Plot
benthic_overall_plot <- ggplot(
  benthic_rel,
  aes(
    x = "Overall",
    y = RelAbundance,
    fill = Benthic.Name
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(
      x = "Overall",
      y = 1.05,
      label = N_quadrats
    ),
    inherit.aes = FALSE,
    fontface = "plain",
    size = 4
  ) +
  scale_fill_manual(
  values = color_mapping,
  labels = benthic_labels
)+
  scale_y_continuous(
    breaks = seq(0, 1, by = 0.25),
    labels = scales::percent_format(accuracy = 1)
  ) +
  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  labs(
    x = NULL,
    y = "Percent Cover",
    fill = "Benthic Type"
  ) +
  theme_bw() +
  theme(legend.position = "none")


benthic_overall_plot
```

Visualize Benthic Cover by Site

```{r}

# Calculate relative abundance by Shore and Site
benthic_rel_site <- data_benthic %>%
  count(Shore, Site, Benthic.Group, Benthic.Name, name = "Count") %>%
  group_by(Shore, Site) %>%
  mutate(RelAbundance = Count / sum(Count)) %>%
  ungroup() %>%
  arrange(Benthic.Group, desc(RelAbundance)) %>%
  mutate(
    Site = factor(Site, levels = site_order),
    Benthic.Name = factor(Benthic.Name, levels = names(color_mapping))
  )

write.csv(benthic_rel_site, "Output/data_benthic_relative_abundance.csv")


# Number of unique quadrats per Shore and Site
site_n <- data_benthic %>%
  distinct(Shore, Site, Transect, Quadrat) %>%
  count(Shore, Site, name = "N_quadrats") %>%
  mutate(
    Site = factor(Site, levels = site_order)
  )

benthic_site_plot <- ggplot(
  benthic_rel_site,
  aes(
    x = Site,
    y = RelAbundance,
    fill = Benthic.Name
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    data = site_n,
    aes(
      x = Site,
      y = 1.03,
      label = N_quadrats
    ),
    inherit.aes = FALSE,
    fontface = "plain",
    size = 2
  ) +
  facet_grid(~ Shore, scales = "free_x", space = "free_x") +
  scale_fill_manual(
  values = color_mapping,
  labels = benthic_labels
)+
  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
)+
  labs(
    x = "Site",
    y = "Percent Cover",
  fill = NULL
  ) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "gray", color = "black"),
    strip.text = element_text(color = "black", face = "bold"),
    legend.text.align = 0,
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 6),
    legend.key.height = unit(0.3, "cm")
  )
benthic_site_plot
```

Calculate relative abundance only for Corals overall

```{r}

data_benthic_coral <- data_benthic %>%
  filter(Benthic.Group=="Hard_Coral")

# Calculate relative abundance
benthic_coral_rel <- data_benthic_coral %>%
  count(Benthic.Group, Benthic.Name, name = "Count") %>%
  mutate(RelAbundance = Count / sum(Count)) %>%
  arrange(Benthic.Group, desc(RelAbundance)) %>%
  mutate(
    Benthic.Name = factor(Benthic.Name, levels = unique(Benthic.Name))
  )

# Labels: italicize genus names, keep non-genus labels plain
non_genus_names <- c("Unidentified coral")

coral_labels <- setNames(
  lapply(levels(benthic_coral_rel$Benthic.Name), function(x) {
    if (x %in% non_genus_names) {
      x
    } else {
      bquote(italic(.(x)))
    }
  }),
  levels(benthic_coral_rel$Benthic.Name)
)

# Number of unique quadrats
N_quadrats <- data_benthic %>%
  distinct(Site, Transect, Quadrat) %>%
  nrow()

unique(data_benthic_coral$Benthic.Name)

coral_types <- sort(unique(data_benthic_coral$Benthic.Name))

#map coral colors
coral_color_mapping <- setNames(
  qualitative_hcl(
    n = length(coral_types),
    palette = "Dark 3"
  ),
  coral_types
)

coral_color_mapping

# Plot
hard_coral_overall_plot <- ggplot(
  benthic_coral_rel,
  aes(
    x = "Overall",
    y = RelAbundance,
    fill = Benthic.Name
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(
      x = "Overall",
      y = 1.05,
      label = N_quadrats
    ),
    inherit.aes = FALSE,
    fontface = "plain",
    size = 4
  ) +
  scale_fill_manual(
  values = coral_color_mapping,
  labels = coral_labels
)+
  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
)+
  labs(
    x = NULL,
    y = "Percent Cover",
    fill = "Hard Coral Type"
  ) +
  theme_bw()+
  theme(legend.position = "none")
hard_coral_overall_plot
```

Calculate relative abundance only for Corals by Site

```{r}
# Calculate Coral relative abundance by Site
benthic_coral_rel_site <- data_benthic_coral %>%
  count(Shore, Site, Benthic.Name, name = "Count") %>%
  group_by(Shore, Site) %>%
  mutate(RelAbundance = Count / sum(Count)) %>%
  ungroup() %>%
  arrange(desc(RelAbundance)) %>%
  mutate(
    Site = factor(Site, levels = site_order),
    Benthic.Name = factor(Benthic.Name, levels = coral_types)
  )

# Number of unique quadrats per Site
site_n_coral <- data_benthic_coral %>%
  distinct(Shore, Site, Transect, Quadrat) %>%
  count(Shore, Site, name = "N_quadrats") %>%
  mutate(Site = factor(Site, levels = site_order))

# Plot
hard_coral_site_plot <- ggplot(
  benthic_coral_rel_site,
  aes(
    x = Site,
    y = RelAbundance,
    fill = Benthic.Name
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    data = site_n,
    aes(
      x = Site,
      y = 1.03,
      label = N_quadrats
    ),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 2
  ) +
  geom_text(
    data = site_n,
    aes(
      x = Site,
      y = 1.03,
      label = N_quadrats
    ),
    inherit.aes = FALSE,
    fontface = "plain",
    size = 2
  ) +
  facet_grid(~ Shore, scales = "free_x", space = "free_x") +
  scale_fill_manual(
  values = coral_color_mapping,
  labels = coral_labels
)+
  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
)+
  labs(
    x = "Site",
    y = "Percent Cover",
  fill = NULL) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "gray", color = "black"),
    strip.text = element_text(color = "black", face = "bold"),
    legend.text.align = 0,
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 6),
    legend.key.height = unit(0.3, "cm")
  )
hard_coral_site_plot
```

Benthic community PERMANOVA and NMDS
```{r}
# Build transect-level community matrix (rows = transects, cols = benthic types)
benthic_comm <- data_benthic |>
  count(Shore, Site, Transect, Benthic.Name, name = "N") |>
  pivot_wider(names_from = Benthic.Name, values_from = N, values_fill = 0)

meta_b    <- benthic_comm |> select(Shore, Site, Transect)
comm_mat_b <- benthic_comm |> select(-Shore, -Site, -Transect)

# Bray-Curtis dissimilarity
dist_b <- vegdist(comm_mat_b, method = "bray")

# --- Site-level PERMANOVA ---
set.seed(1234)
site_permanova_b <- adonis2(dist_b ~ Site, data = meta_b, permutations = 9999)
site_permanova_b

# Dispersion (homogeneity of within-site variance)
disp_site_b      <- betadisper(dist_b, meta_b$Site)
disp_site_anova  <- anova(disp_site_b)
disp_site_perm   <- permutest(disp_site_b, permutations = 9999)
cat("\nSite dispersion ANOVA:\n"); print(disp_site_anova)
cat("\nSite dispersion permutation test:\n"); print(disp_site_perm)

# --- NMDS ---
set.seed(1234)
nmds_b <- metaMDS(comm_mat_b, distance = "bray", k = 2, trymax = 100)
cat("Stress:", round(nmds_b$stress, 3), "\n")
#check stress values

#extract scores
scores_b <- as.data.frame(scores(nmds_b, display = "sites")) |>
  bind_cols(meta_b)

#calculate centroids
centroids_b <- scores_b |>
  group_by(Shore) |>
  summarise(NMDS1 = mean(NMDS1), NMDS2 = mean(NMDS2), .groups = "drop")

#calculate hulls
hulls_b <- scores_b |>
  group_by(Shore) |>
  filter(n() >= 3) |>
  dplyr::slice(chull(NMDS1, NMDS2)) |>
  ungroup()

#benthic types by NMDS loading magnitude
sp_scores_b <- as.data.frame(scores(nmds_b, display = "species")) |>
  rownames_to_column("Benthic") |>
  mutate(mag = sqrt(NMDS1^2 + NMDS2^2)) |>
  arrange(desc(mag)) |>
  head(20)

shore_cols_b <- c(North = "turquoise3", East = "palegreen3", West = "plum3")

site_shapes_b <- c(
  S2=21, S3=22, S4=23, S5=24,
  S6=25, S9=21, S11=22,
  S12=23, S14=24, S15=25, S18=21
)

#plot NMDS of benthic cover
benthic_nmds_plot <- ggplot(scores_b, aes(NMDS1, NMDS2)) +
  geom_polygon(data = hulls_b, aes(fill = Shore, group = Shore),
               alpha = 0.25, color = NA) +
  geom_point(aes(fill = Shore, shape = Site),
             color = "black", size = 3, stroke = 0.5) +
  geom_text(data = centroids_b, aes(label = Shore, color = Shore),
            fontface = "bold", size = 4) +
  geom_segment(data = sp_scores_b,
               aes(x = 0, y = 0, xend = NMDS1 * 0.6, yend = NMDS2 * 0.6),
               arrow = arrow(length = unit(0.2, "cm")),
               color = "gray40", linewidth = 0.5) +
  geom_text(data = sp_scores_b,
            aes(x = NMDS1 * 0.65, y = NMDS2 * 0.65, label = Benthic),
            size = 2.8, color = "gray30", fontface = "italic") +
  scale_fill_manual(values = shore_cols_b) +
  scale_color_manual(values = shore_cols_b) +
  scale_shape_manual(values = site_shapes_b) +
  annotate("text", x = Inf, y = Inf,
           label = {
             fmt_p <- function(p) {
               if (is.na(p))      return("NA")
               if (p < 0.0001)    "< 0.0001"
               else if (p < 0.001) "< 0.001"
               else if (p < 0.01)  "< 0.01"
               else if (p < 0.05)  "< 0.05"
               else                as.character(round(p, 3))
             }
             perm_F  <- round(site_permanova_b$F[1], 2)
             perm_R2 <- round(site_permanova_b$R2[1], 2)
             perm_p  <- fmt_p(site_permanova_b[["Pr(>F)"]][1])
             disp_p  <- fmt_p(disp_site_perm$tab[["Pr(>F)"]][1])
             paste0(
               "Site PERMANOVA:  F = ", perm_F,
               ", R\u00b2 = ", perm_R2,
               ", p = ", perm_p, "\n",
               "Site dispersion: p = ", disp_p,
               "  |  Stress = ", round(nmds_b$stress, 3)
             )
           },
           hjust = 1.05, vjust = 1.2, size = 2.8, color = "gray30") +
  labs(x = "NMDS1", y = "NMDS2",
       title = "Benthic community composition by shore",
       fill = "Shore", shape = "Site") +
  guides(color = "none",
         shape = guide_legend(override.aes = list(fill = "gray70"))) +
  theme_bw()

benthic_nmds_plot

ggsave(
  "Output/benthic_nmds_plot.jpg",
  benthic_nmds_plot,
  device = "jpg", width = 8, height = 6, units = "in"
)
```

Visualize the size and species data

```{r}

#Set the Site Order
data$Site <- factor(
  data$Site,
  levels = c(
    "S2","S3","S4","S5","S6","S9","S11","S12","S14","S15","S18"
  )
)

#Set the Species Order
data$Species <- factor(
  data$Species,
  levels = c( "P. acuta", "P. effusa","P. grandis","P. meandrina","P. tuahiniensis", "P. verrucosa"
  )
)

#Calculate relative abundance by species and site
rel_abund <- data %>%
  count(Shore, Site, Species) %>%
  group_by(Shore, Site) %>%
  mutate(RelAbundance = n / sum(n)) %>%
  ungroup()

#Count sample size
site_n <- data %>%
  count(Shore, Site, name = "N")

pocillopora_site_plot <- ggplot(rel_abund,
       aes(x = Site,
           y = RelAbundance,
           fill = Species)) +
  geom_col() +

  geom_text(
    data = site_n,
    aes(x = Site, y = 1.03, label = N),
        inherit.aes = FALSE,
    fontface = "plain",
    size = 2
  ) +
  labs(
  x = "Site",
  y = "Relative Abundance",
  fill = NULL)+

  facet_grid(~ Shore, scales = "free_x", space = "free_x") +

  scale_fill_manual(
    values = c(
      "P. tuahiniensis" = "#D55E00",
      "P. meandrina"    = "#0072B2",
      "P. verrucosa"    = "#E69F00",
      "P. grandis"      = "#56B4E9",
      "P. effusa"       = "#009E73",
      "P. acuta"        = "#e63946"
    ),
    labels = c(
      "P. tuahiniensis" = expression(italic("P. tuahiniensis")),
      "P. meandrina"    = expression(italic("P. meandrina")),
      "P. verrucosa"    = expression(italic("P. verrucosa")),
      "P. grandis"      = expression(italic("P. grandis")),
      "P. effusa"       = expression(italic("P. effusa")),
      "P. acuta"        = expression(italic("P. acuta"))
    )
  ) +

  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
)+
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "gray", color = "black"),
    strip.text = element_text(color = "black", face = "bold"),
    legend.text.align = 0,
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 6),
    legend.key.height = unit(0.4, "cm")
  )
pocillopora_site_plot

```

PERMANOVA test of relative abundance by Site

```{r}
#calculate the community by Shore, Site, Transect, Species
comm <- data %>%
  count(Shore, Site, Transect, Species) %>%
  group_by(Shore, Site, Transect) %>%
  mutate(RelAbundance = n / sum(n)) %>%
  ungroup() %>%
  select(Shore, Site, Transect, Species, RelAbundance) %>%
  pivot_wider(
    names_from = Species,
    values_from = RelAbundance,
    values_fill = 0
  )

#set the data information
meta <- comm %>%
  select(Shore, Site, Transect)

#identify the community matrix
comm_mat <- comm %>%
  select(-Shore, -Site, -Transect)

#calculate the bray curtis distance
dist_bc <- vegdist(comm_mat, method = "bray")

# run the PERMANOVA by site
set.seed(1234)
site.per <- adonis2(
  dist_bc ~ Site,
  data = meta,
  permutations = 9999
)
site.per

#calculate beta dispersion
disp <- betadisper(dist_bc, meta$Site)
anova(disp)
permutest(disp)

# run the NMDS
set.seed(1234)
nmds <- metaMDS(
  comm_mat,
  distance = "bray",
  k = 2,
  trymax = 100
)

#extract the scores
scores_df <- as.data.frame(scores(nmds, display = "sites"))

scores_df$Site <- meta$Site
scores_df$Transect <- meta$Transect
scores_df$Shore <- meta$Shore

#Calculate the centroids
centroids <- scores_df %>%
  group_by(Shore, Site) %>%
  summarise(
    NMDS1 = mean(NMDS1),
    NMDS2 = mean(NMDS2),
    .groups = "drop"
  )

site_cols <- c(
  # North - blues
  "S2"  = "turquoise2",
  "S3"  = "turquoise2",
  "S4"  = "turquoise2",
  "S5"  = "turquoise2",

  # East - greens
  "S6"  = "palegreen1",
  "S9"  = "palegreen1",
  "S11" = "palegreen1",

  # West - oranges/reds
  "S12"  = "plum2",
  "S14" = "plum2",
  "S15" = "plum2",
  "S18" = "plum2"
)


#plot
ggplot(scores_df,
       aes(NMDS1, NMDS2,
           color = Site,
           shape = Site)) +

  geom_polygon(
    data = scores_df %>%
      group_by(Site) %>%
      slice(chull(NMDS1, NMDS2)),
    aes(fill = Site),
    alpha = 0.3,
    color = NA
  ) +

  geom_point(
    size = 2,
    alpha = 0.8
  ) +

geom_text(
  data = centroids,
  aes(label = Site),
  color = "black",
  fontface = "bold",
  size = 3,
  nudge_y = 0.02
)+

  scale_color_manual(values = site_cols) +
  scale_fill_manual(values = site_cols) +

  scale_shape_manual(
    values = c(
      S2 = 1,  # solid circle
      S3  = 2,  # solid triangle
      S4  = 3,   # solid square
      S5 = 4,  # solid circle
      S6  = 5,  # solid triangle
      S9  = 6,   # solid square
      S11 = 7,  # solid circle
      S12  = 8,  # solid triangle
      S14  = 9,   # solid square
      S15  = 10,   # solid square
      S18  = 11   # solid square
    )
  ) +

  theme_bw()
```

Plot total Pocillopora with all sites together

```{r}
# Calculate overall relative abundance
rel_abund_overall <- data %>%
  count(Species) %>%
  mutate(RelAbundance = n / sum(n))

# Total sample size
N_total <- nrow(data)

pocillopora_overall_plot <- ggplot(rel_abund_overall,
       aes(x = "Overall",
           y = RelAbundance,
           fill = Species)) +
  geom_col(width=0.7) +

  geom_text(
    aes(x = "Overall",
        y = 1.05,
        label = N_total),
    inherit.aes = FALSE,
    fontface = "plain",
    size = 3.5
  ) +

  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Species"
  ) +

  scale_fill_manual(
    values = c(
      "P. tuahiniensis" = "#D55E00",
      "P. meandrina"    = "#0072B2",
      "P. verrucosa"    = "#E69F00",
      "P. grandis"      = "#56B4E9",
      "P. effusa"       = "#009E73",
      "P. acuta"        = "#e63946"
    ),
    labels = c(
      "P. tuahiniensis" = expression(italic("P. tuahiniensis")),
      "P. meandrina"    = expression(italic("P. meandrina")),
      "P. verrucosa"    = expression(italic("P. verrucosa")),
      "P. grandis"      = expression(italic("P. grandis")),
      "P. effusa"       = expression(italic("P. effusa")),
      "P. acuta"        = expression(italic("P. acuta"))
    )
  ) +

  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
)+
  theme_bw()+
  theme(legend.position = "none")
pocillopora_overall_plot

```

Plot multipanel figures

```{r}
# plot objects
Fig1 <- map_plot + labs(tag = "A")
Fig2 <- benthic_overall_plot + labs(tag = "B")
Fig3 <- hard_coral_overall_plot + labs(tag = "D")
Fig4 <- pocillopora_overall_plot + labs(tag = "F")

Fig5 <- benthic_site_plot + labs(tag = "C")
Fig6 <- hard_coral_site_plot + labs(tag = "E")
Fig7 <- pocillopora_site_plot + labs(tag = "G")

row1 <- Fig1

row2 <- (Fig2 + Fig5) +
  plot_layout(widths = c(0.8, 1.8))

row3 <- (Fig3 + Fig6) +
  plot_layout(widths = c(0.8, 1.8))

row4 <- (Fig4 + Fig7) +
  plot_layout(widths = c(0.8, 1.8))

multi_panel_fig <-
  row1 /
  row2 /
  row3 /
  row4 +
  plot_layout(
    heights = c(1.8, 1, 1, 1)
  ) &
  theme(
    plot.tag = element_text(
      face = "bold",
      size = 18
    ),
    plot.tag.position = c(0.02, 0.98)
  )

# ggsave(
#   "Output/multi_panel_fig.pdf",
#   multi_panel_fig,
#   width = 12,
#   height = 10,
#   units = "in",
#   device = cairo_pdf,
#   bg = "white"
# )
# 
# ggsave(
#   "Output/multi_panel_fig.jpg",
#   multi_panel_fig,
#   width = 12,
#   height = 10,
#   units = "in",
#   bg = "white"
# )


```

Relative Abundance by Site and Size Class

```{r}

#Calculate realtive abundance
rel_abund <- data %>%
  count(Shore, Site, SizeClass, Species) %>%
  group_by(Shore, Site, SizeClass) %>%
  mutate(RelAbundance = n / sum(n)) %>%
  ungroup()

site_n <- data %>%
  count(Shore, Site, SizeClass, name = "N")


pocillopora_size_plot <- ggplot(rel_abund,
       aes(x = Site,
           y = RelAbundance,
           fill = Species)) +

  geom_col() +

  geom_text(
    data = site_n,
    aes(x = Site, y = 1.03, label = N),
    inherit.aes = FALSE,
    size = 2.5
  ) +

  scale_fill_manual(
    values = c(
      "P. tuahiniensis" = "#D55E00",
      "P. meandrina"    = "#0072B2",
      "P. verrucosa"    = "#E69F00",
      "P. grandis"      = "#56B4E9",
      "P. effusa"       = "#009E73",
      "P. acuta"        = "#e63946"
    ),
    labels = c(
      "P. tuahiniensis" = expression(italic("P. tuahiniensis")),
      "P. meandrina"    = expression(italic("P. meandrina")),
      "P. verrucosa"    = expression(italic("P. verrucosa")),
      "P. grandis"      = expression(italic("P. grandis")),
      "P. effusa"       = expression(italic("P. effusa")),
      "P. acuta"        = expression(italic("P. acuta"))
    )
  ) +

  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1))+

  labs(
    x = "Site",
    y = "Percent Cover",
    fill = NULL
  ) +

  theme_bw() +
  theme(
    strip.background = element_rect(fill = "gray", color = "black"),
    strip.text = element_text(color = "black", face = "bold"),
    legend.text.align = 0,
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

pocillopora_size_plot.2 <- pocillopora_size_plot +
  facet_grid2(
    SizeClass ~ Shore,
    scales = "free_x",
    space = "free_x",
    strip = strip_themed(
      background_x = elem_list_rect(
  fill = c(
    alpha("turquoise2", 0.4),  # North
    alpha("palegreen1", 0.4),  # East
    alpha("plum2", 0.4)        # West
  ),
        color = "black"
      ),
      text_x = elem_list_text(
        face = "bold",
        color = "black"
      )
    )
  )
pocillopora_size_plot.2
```

PERMANOVA, dispersion test, NMDS, and plotting

```{r}
# Shore colors
shore_cols <- c(
  North = "turquoise2",
  East  = "palegreen1",
  West  = "plum2"
)

# Site symbols: fillable shapes so shore color appears inside symbol
site_shapes <- c(
  S2  = 21,
  S3  = 22,
  S4  = 23,
  S5  = 24,
  S6  = 25,
  S9  = 21,
  S11 = 22,
  S12 = 23,
  S14 = 24,
  S15 = 25,
  S18 = 21
)

# Function that runs PERMANOVA, dispersion test, NMDS, and generates a plot
run_sizeclass_nmds <- function(size_class_name) {

  data_sub <- data %>%
    filter(SizeClass == size_class_name)

  comm <- data_sub %>%
    count(Shore, Site, Transect, Species) %>%
    group_by(Shore, Site, Transect) %>%
    mutate(RelAbundance = n / sum(n)) %>%
    ungroup() %>%
    select(Shore, Site, Transect, Species, RelAbundance) %>%
    pivot_wider(
      names_from = Species,
      values_from = RelAbundance,
      values_fill = 0
    )

  meta <- comm %>%
    select(Shore, Site, Transect)

  comm_mat <- comm %>%
    select(-Shore, -Site, -Transect)

  dist_bc <- vegdist(comm_mat, method = "bray")

  set.seed(1234)
  site_per <- adonis2(
    dist_bc ~ Site,
    data = meta,
    permutations = 9999
  )

  disp <- betadisper(dist_bc, meta$Site)
  disp_anova <- anova(disp)
  set.seed(1234)
  disp_perm <- permutest(disp, permutations = 9999)

  set.seed(1234)
  nmds <- metaMDS(
    comm_mat,
    distance = "bray",
    k = 2,
    trymax = 100
  )

  per_tab <- as.data.frame(site_per)

  site_p <- per_tab["Site", "Pr(>F)"]

  stat_label <- paste0(
    "p = ", signif(site_p, 3),
    "\nStress = ", round(nmds$stress, 3)
  )

  scores_df <- as.data.frame(scores(nmds, display = "sites"))

  scores_df$Shore <- meta$Shore
  scores_df$Site <- meta$Site
  scores_df$Transect <- meta$Transect
  scores_df$SizeClass <- size_class_name

  scores_df$Site <- droplevels(scores_df$Site)
  scores_df$Shore <- droplevels(scores_df$Shore)

  sites_in_plot <- levels(scores_df$Site)

  site_shore_lookup <- scores_df %>%
    distinct(Site, Shore) %>%
    mutate(Site = as.character(Site)) %>%
    filter(Site %in% sites_in_plot) %>%
    arrange(match(Site, sites_in_plot))

  centroids <- scores_df %>%
    group_by(Shore, Site) %>%
    summarise(
      NMDS1 = mean(NMDS1),
      NMDS2 = mean(NMDS2),
      .groups = "drop"
    )

  hulls <- scores_df %>%
    group_by(Site) %>%
    filter(n() >= 3) %>%
    slice(chull(NMDS1, NMDS2))

  p <- ggplot(
    scores_df,
    aes(
      NMDS1,
      NMDS2,
      shape = Site
    )
  ) +

    geom_polygon(
      data = hulls,
      aes(fill = Shore, group = Site),
      alpha = 0.3,
      color = NA
    ) +

    geom_point(
      aes(fill = Shore),
      color = "black",
      size = 2,
      stroke = 0.6,
      alpha = 0.9
    ) +

    geom_text(
      data = centroids,
      aes(label = Site),
      color = "black",
      fontface = "bold",
      size = 2
    ) +

    annotate(
      "text",
      x = Inf,
      y = Inf,
      label = stat_label,
      hjust = 1.05,
      vjust = 1.1,
      size = 3
    ) +

    scale_fill_manual(values = shore_cols) +

    scale_shape_manual(
      values = site_shapes,
      breaks = sites_in_plot
    ) +

    guides(
  fill = "none",
  shape = guide_legend(
    title = NULL,
    nrow = 2,
    byrow = TRUE,
    override.aes = list(
      fill = shore_cols[as.character(site_shore_lookup$Shore)],
      color = "black",
      size = 3,
      stroke = 0.4,
      alpha = 1
    )
  )
) +

    labs(
      x = "NMDS1",
      y = "NMDS2",
      shape = "Site"
    ) +

    theme_bw()

  list(
    size_class = size_class_name,
    permanova = site_per,
    dispersion_anova = disp_anova,
    dispersion_permutest = disp_perm,
    nmds = nmds,
    scores = scores_df,
    centroids = centroids,
    plot = p
  )
}

# Run for all size classes
size_results <- map(levels(data$SizeClass), run_sizeclass_nmds)
names(size_results) <- levels(data$SizeClass)

# Summary table
size_summary <- map_dfr(size_results, function(x) {

  per <- as.data.frame(x$permanova)

  tibble(
    SizeClass = x$size_class,
    Df        = per["Site", "Df"],
    F         = per["Site", "F"],
    p.value   = per["Site", "Pr(>F)"],
    Stress    = x$nmds$stress
  )
})

size_summary

# Extract plots
pocillopora_NMDS_02 <- size_results[["0-2cm"]]$plot +
  labs(x = NULL) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

pocillopora_NMDS_25 <- size_results[[">2cm-5cm"]]$plot +
  labs(x = NULL) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

pocillopora_NMDS_514 <- size_results[[">5cm-14cm"]]$plot +
  labs(x = NULL) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

pocillopora_NMDS_14up <- size_results[[">14cm"]]$plot +
  labs(x = "NMDS1") +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(size = 9),
    legend.key.height = unit(0.5, "cm"),
    legend.key.width = unit(0.75, "cm")
  )
```

Multipanel plot

```{r}
# Plot objects
Fig8 <- pocillopora_size_plot.2 + labs(tag = "A")
Fig9 <- pocillopora_NMDS_02 + labs(tag = "B")
Fig10 <- pocillopora_NMDS_25 + labs(tag = "C")
Fig11 <- pocillopora_NMDS_514 + labs(tag = "D")
Fig12 <- pocillopora_NMDS_14up + labs(tag = "E")

right_column <-
  Fig9 /
  Fig10 /
  Fig11 /
  Fig12

multi_panel_nmds <-
  Fig8 | right_column +
  plot_layout(
    widths = c(1.5, 1),
    heights = c(1, 1, 1, 1)
  )

# ggsave(
#   "Output/multi_panel_nmds.pdf",
#   multi_panel_nmds,
#   width = 10,
#   height = 14,
#   units = "in",
#   device = cairo_pdf,
#   bg = "white"
# )
# 
# ggsave(
#   "Output/multi_panel_nmds.jpg",
#   multi_panel_nmds,
#   width = 10,
#   height = 14,
#   units = "in",
#   bg = "white"
# )

```

