---
title: "Relative_abundance"
output: html_document
---
This script aims to analyse the relative abundance of Pocillopora around the island of Mo’orea at 12 sites. This script also aims to characterize the benthos at each site in order to better understand the biodiversity in which the Pocillopora community lives.

Q1: What does the benthic cover of the lagoon around the island look like?
Q2: What corals are found on the lagoon reefs of Mo’orea?
Q3: How are Pocillopora species distributed around the lagoon?
H0: Benthic cover, coral cover, and Pocillopora relative abundance differ by site.

#Install and load packages
```{r}
library("ggplot2")
library("vegan")
library("tidyverse")
library("purrr")
library("ggsn")
library("colorspace")
library("patchwork")
library("ggmap")
library("ggh4x")
```

Load Data Files for Analysis
```{r}
#read in species id data
data_haplotype <- read.csv("Data/Haplotype_data_modified_duplication.csv")

#read in colony diameter data
data_size <- read.csv("Data/POC_RA_Size.csv")

#read in benthic quadrat data
data_benthic <- read.csv("Data/20260226_Quadrat_Benthos_RA.csv")
```

Data Cleaning species and diameter
```{r}
#filter to only data from the first 3 transects
data_size <- data_size %>%
  filter(Transect %in% c("T1", "T2", "T3"))

# Remove uninformative columns ("Sanger.Plate.Name","Sequence.ID","Plate.Well", "gDNA.Extraction.Plate") and filter out NA rows
data_species <- data_haplotype %>%
  select(-c("Sanger.Plate.Name","Sequence.ID","Plate.Well", "gDNA.Extraction.Plate", "Notes")) %>% na.omit()

# Remove uninformative columns ("Diameter.1","Diameter.2","Diameter.3","Note") and filter out NA rows and DIV/0! rows and check if diameter is numeric
data_diameters <- data_size %>%
  select(-c("Diameter.1","Diameter.2","Diameter.3","Note")) %>%
  na.omit() %>%
  filter(Diameter.mean!="#DIV/0!") 

str(data_diameters)
data_diameters$Diameter.mean <- as.numeric(data_diameters$Diameter.mean)
str(data_diameters)

```

Quality Control of Sample.IDs
```{r}
# Check species and diameter datasets by Sample.ID to see if there are any ID problems

# Check to see they are all distinct
dup_species <- data_species %>% count(Sample.ID, sort = TRUE) %>% filter(n > 1)
dup_species$Sample.ID <- as.numeric(dup_species$Sample.ID)
  
#generate a list of duplicated IDs
dup_rows_species <- data_species %>%
  filter(Sample.ID %in% dup_species$Sample.ID) %>%
  arrange(Sample.ID)

#save a file of problematic Sample.ID to QC
write.csv(dup_rows_species, "Output/dup_rows_species.csv")

# Check to see they are all distinct
dup_diameters <-data_diameters %>% count(Sample.ID, sort = TRUE) %>% filter(n > 1)
dup_diameters$Sample.ID <- as.numeric(dup_diameters$Sample.ID)

#generate a list of duplicated IDs
dup_rows_size <- data_diameters %>%
  filter(Sample.ID %in% dup_diameters$Sample.ID) %>%
  arrange(Sample.ID)

#save a file of problematic Sample.ID to QC
write.csv(dup_rows_size, "Output/dup_rows_size.csv")

# Check to see if all Sample.ID are all numeric
bad_ids_species <- data_species$Sample.ID[
  is.na(suppressWarnings(
    as.numeric(as.character(data_species$Sample.ID))
  ))
]
bad_ids_species

#An issues come from 2 samples because _BIS is present at the end of the Sample.ID
#For these samples "_BIS" means #.2 in French, essentially a duplicate name
#We do not have pictures of these two _BIS corals so they are going to be excluded from the dataset

data_species <- data_species %>%
  filter(!grepl("^[0-9]+_BIS$", Sample.ID)) %>% #filter out _BIS samples
  mutate(Sample.ID = as.numeric(as.character(Sample.ID))) # change the sample.ID to numeric
  
# Check to see if all Sample.ID are all numeric
bad_ids_species <- data_species$Sample.ID[
  is.na(suppressWarnings(
    as.numeric(as.character(data_species$Sample.ID))
  ))
]
bad_ids_species
#There are only numeric IDs in species data

# Check to see if all Sample.ID are all numeric
bad_ids_diameters <- data_diameters$Sample.ID[
  is.na(suppressWarnings(
    as.numeric(as.character(data_diameters$Sample.ID))
  ))
]
bad_ids_diameters 
#There are only numeric IDs in diameter data

# Remove remaining duplicates if QC and manual check of the photos and notebooks do not reconcile and you cannot confirm the sample ID
data_species <- data_species %>%
  filter(!Sample.ID %in% dup_species$Sample.ID)

data_diameters <- data_diameters %>% 
  filter(!Sample.ID %in% dup_diameters$Sample.ID)

```

Sanity checks for data quality
```{r}
# This section verifies the integrity of the datasets before
# any analysis. Each check is documented for reproducibility
# and transparency for reviewers.

# --- QC 1: Dataset Dimensions -------------------------------
# Verify the number of rows and columns after cleaning
cat("data_species dimensions:  ", dim(data_species),  "\n")
cat("data_diameters dimensions:", dim(data_diameters), "\n")
#when all species IDs are complete we would expect the same number of rows in both dataframes
#We have ~300 more species than we do sizes, which is likely due to the species coming from transects 4,5,6 and we are not able to filter them by transect becasue the file did not have metadata within it


# --- QC 2: Dataset Structure --------------------------------
# Verify column names and types are as expected
#Species should be character
#Sample.ID should be numeric
#Diameter.mean should be numeric
str(data_species)
str(data_diameters)


# --- QC 3: Missing Values -----------------------------------
# Confirm no NAs remain after na.omit in cleaning step
#results from this should be all 0's
cat("\nMissing values in data_species:\n")
colSums(is.na(data_species))

cat("\nMissing values in data_diameters:\n")
colSums(is.na(data_diameters))


# --- QC 4: Duplicate Sample.IDs -----------------------------
# Confirm no duplicates remain after deduplication step
cat("\nRemaining duplicates in data_species:\n")
data_species  %>% count(Sample.ID) %>% filter(n > 1)

cat("\nRemaining duplicates in data_diameters:\n")
data_diameters %>% count(Sample.ID) %>% filter(n > 1)

# --- QC 5: Value Range & Reality Check ------------------------
# Check for impossible or unexpected values in key variables
#this should be numeric, not character
cat("\nSummary of data_diameters:\n")
range(data_diameters$Diameter.mean)
#this should range from >0-100, where 100 indicates everything >14cm

cat("\nUnique sites found in data_diameters:\n")
unique(data_diameters$Site)

cat("\nUnique transects found in data_diameters:\n")
unique(data_diameters$Transect)

cat("\nUnique species found in data_species:\n")
#should be "P. grandis" "P. meandrina" "P. verrucosa" "P. effusa" "P. tuahiniensis" "P. acuta"  
unique(data_species$Species)

cat("\nUnique haplotypes found in data_species:\n")
#should be "Haplotype 1a" "Haplotype 3"  "Haplotype 2"  "Haplotype 11" "Haplotype 10" "Haplotype 8a" "Haplotype 5a"
unique(data_species$Haplotype)


```

Join the species and size datasets together for Size Class specific analyses
```{r}
# --- QC 6: Join Integrity -----------------------------------
# Identify Sample.IDs present in one dataset but not the other
# These samples will be excluded from the merged dataset
lost_from_species <- anti_join(data_species,
  data_diameters,
  by = "Sample.ID"
)

lost_from_diameters <- anti_join(
  data_diameters,
  data_species,
  by = "Sample.ID"
)

# Store excluded Sample.IDs for traceability
excluded_from_species   <- lost_from_species  %>% pull(Sample.ID)
excluded_from_diameters <- lost_from_diameters %>% pull(Sample.ID)

cat("\nSamples excluded from data_species (no size match):",   length(excluded_from_species),   "\n")
cat("Samples excluded from data_diameters (no species match):", length(excluded_from_diameters), "\n")

# --- Remove unmatched samples from each dataset -------------
data_species <- data_species %>%
  filter(!Sample.ID %in% excluded_from_species)
dim(data_species)

data_diameters <- data_diameters %>%
  filter(!Sample.ID %in% excluded_from_diameters)
dim(data_diameters)

# --- Rebuild merged dataset after exclusions ----------------
data <- data_species %>%
  inner_join(data_diameters, by = "Sample.ID")

cat("\nFinal merged dataset dimensions after exclusions:", dim(data), "\n")

# --- Confirm no coercion warnings remain --------------------
cat("\nFinal merged dataset dimensions after exclusions:", dim(data), "\n")
cat("Missing values in final merged dataset:\n")
colSums(is.na(data))

# Check all unique species names
sort(unique(data$Species))
sort(unique(data$Haplotype))

sort(unique(data$Site))
sort(unique(data$Transect))

# Cross-check: how many transects per site?
data  %>%
  group_by(Site)  %>%
  summarise(n_transects = n_distinct(Transect),
            n_samples   = n())


# --- Assign colonies to Size Class ---------------
# Colonies recorded as 100cm represent colonies whose diameter
# could not be measured in the field (> 100cm). These are
# reclassified as a distinct ">100cm" size class rather than
# being removed, as they represent real biological observations.

data <- data %>%
  mutate(
    SizeClass = case_when(
      Diameter.mean <= 2 ~ "0-2cm",
      Diameter.mean > 2 & Diameter.mean <= 5 ~ ">2cm-5cm",
      Diameter.mean > 5 & Diameter.mean <= 14 ~ ">5cm-14cm",
      Diameter.mean > 14 ~ ">14cm",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    SizeClass = factor(
      SizeClass,
      levels = c("0-2cm", ">2cm-5cm", ">5cm-14cm", ">14cm")
    )
  )

#because some NA have been find into the dataset, we will remove them to avoid any trouble during the stats analysis 
# Remove rows with NA in SizeClass
data <- data %>%
  filter(!is.na(SizeClass))

# Confirm no NAs remain
cat("Remaining NAs in SizeClass:", sum(is.na(data$SizeClass)), "\n")
cat("Remaining NAs in Site:", sum(is.na(data$Site)), "\n")
cat("Remaining NAs in Transect:", sum(is.na(data$Transect)), "\n")
cat("Final dataset dimensions:", dim(data), "\n")
```


Add shore category based on site information
```{r}
#for the rest of the analysis we add the shore for species distribution analysis
# Manually assign each site to its corresponding shore
# based on the geographic position around Mo'orea island
# North: sites S1, S2, S3, S4, S5
# East:  sites S6, S9, S11
# West:  sites S12, S14, S15, S18

data <- data %>%
  mutate(Shore = case_when(
    Site %in% c("S1", "S2", "S3", "S4", "S5")       ~ "North",
    Site %in% c("S6", "S9", "S11")         ~ "East",
    Site %in% c("S12", "S14", "S15", "S18")   ~ "West",
    .default = NA_character_
  )) %>%
  mutate(Shore = factor(Shore, levels = c("North", "East", "West")))

# Verify shore assignment
data %>%
  group_by(Shore, Site) %>%
  summarise(n_samples = n(), .groups = "drop") %>%
  arrange(Shore, Site)

# Confirm no unassigned sites
cat("Samples with no shore assigned:", sum(is.na(data$Shore)), "\n")

```
Moorea site map
```{r}
register_google(
  key = Sys.getenv("GOOGLE_MAPS_API_KEY")
)
# location
Moorea<-data.frame(lon = -149.83246425684064, lat = -17.531092816791094)

#Map base
M1<-get_map(Moorea,zoom = 12, maptype = 'satellite')

bbx <- c(left=-149.802,bottom= -17.480,right=-149.805,top=-17.475)
x <- c(bbx["left"], bbx["left"], bbx["right"], bbx["right"])
y <- c(bbx["bottom"], bbx["top"], bbx["top"], bbx["bottom"])
df <- data.frame(x, y)

sitedata <- read.csv("Data/site.coords.csv")
str(sitedata)

labels <- sitedata$Site

shore_bars <- data.frame(
  Shore = c("North", "East", "West"),
  color = c("turquoise2", "palegreen1", "plum2"),

  # edit these coordinates to place bars exactly where you want them
  x    = c(-149.92389316984602, -149.74536533966472, -149.86415501188452),
  xend = c(-149.7767793719456, -149.79188557213965, -149.94215561),
  y    = c(-17.4709289594418, -17.496471002597975, -17.607929750776858),
  yend = c(-17.45799283993501, -17.604330100673128, -17.49303285959368)
)

map_plot <- ggmap(M1) +
  geom_segment(
    data = shore_bars,
    aes(x = x, y = y, xend = xend, yend = yend, color = Shore),
    linewidth = 1,
    lineend = "round"
  ) +
  geom_point(
    data = sitedata,
    aes(x = long, y = lat),
    size = 2,
    color = "yellow"
  ) +
  geom_text(
    data = sitedata,
    aes(x = long, y = lat, label = Site),
    vjust = -1,
    size = 3,
    color = "yellow"
  ) +
  scale_color_manual(
    values = c(
      North = "turquoise2",
      East  = "palegreen1",
      West  = "plum2"
    ),
    guide = "none"
  ) +
  scalebar(
    data = sitedata,
    dist = 2,
    dist_unit = "km",
    transform = TRUE,
    model = "WGS84",
    st.size = 3,
    st.color = "white",
    box.color = "white",
    height = 0.001,
    location = "bottomleft"
  ) +
  xlab("") +
  ylab("")

map_plot
```

Data Cleaning benthic
```{r}

data_benthic <- read.csv("Data/20260226_Quadrat_Benthos_RA.csv") 

data_benthic <- data_benthic %>%
  rename(Benthic.Type = Label.code)

#the naming structure should be Quadrat_Site_Transect.jpg or jpeg
#check to see if there are more than two _ in a Name
data_benthic %>%
  mutate(
    n_underscore = str_count(Name, "_"),
    prefix = str_extract(Name, "^[^_]+_[^_]+_[^_]+")
  ) %>%
  group_by(prefix) %>%
  mutate(prefix_count = n()) %>%
  ungroup() %>%
  filter(n_underscore >= 3) %>%
  select(Name, prefix, prefix_count) %>%
  distinct()
#There are more than two _ in a name for several, but they each have 100 points, which is the number of points counted in CoralNet

data_benthic <- data_benthic %>%
  mutate(Name = sub("\\.JPG$", "", Name, ignore.case = TRUE)) %>%
  mutate(Name = sub("\\.jpeg$", "", Name, ignore.case = TRUE)) %>%
  separate(
    Name,
    into = c("Quadrat", "Site", "Transect"),
    sep = "_",
    remove = TRUE
  )

#filter to only T1-T3
data_benthic <- data_benthic %>%
  filter(Transect %in% c("T1", "T2", "T3")) %>% filter(!Site =="S1")

#check to see if all quadrat, site, and transect numbers exist
unique(data_benthic$Quadrat) 
unique(data_benthic$Site) 
unique(data_benthic$Transect) 

#set site order
site_order <- c(
  "S2","S3","S4","S5","S6","S9","S11","S12","S14","S15","S18"
)

#Add Shore information
data_benthic <- data_benthic %>%
  mutate(Shore = case_when(
    Site %in% c("S1", "S2", "S3", "S4", "S5")       ~ "North",
    Site %in% c("S6", "S9", "S11")         ~ "East",
    Site %in% c("S12", "S14", "S15", "S18")   ~ "West",
    .default = NA_character_
  )) %>%
  mutate(Shore = factor(Shore, levels = c("North", "East", "West")))

unique(data_benthic$Shore)

#count the number of Benthic.Type points by quadrat transect and site
data_benthic_counts <- data_benthic %>%
  count(Site, Transect, Quadrat, name = "Count")
#they should all be 100

#count the number of quadrats per transect and site
quadrat_counts <- data_benthic %>%
  distinct(Site, Transect, Quadrat) %>%
  count(Site, Transect, name = "N_Quadrats")
#they should be a maximum of 26

# Add ecological group for color mapping and filtering
data_benthic <- data_benthic %>%
  mutate(
    Benthic.Group = case_when(
      Benthic.Type %in% c(
        "Acanthastr", "Acropora", "Astreopora", "Cyphastrea",
        "Fungia", "Gardinoser", "Leptastrea", "Leptoseris",
        "Lobophylli", "Montastrea", "Montipora", "Pachyseris",
        "Pavona", "Pocillopo", "Porites", "Psammomora",
        "Tubastrea", "Sandolitha", "Un coral"
      ) ~ "Hard_Coral",

      Benthic.Type %in% c(
        "Macroalgae"
      ) ~ "Algae",
      
      Benthic.Type %in% c(
        "CCA_bare"
      ) ~ "CTB",

      Benthic.Type %in% c(
        "Sand", "Unk", "GiCl", "Millepora"
      ) ~ "Misc",

      TRUE ~ "Other"
    )
  )

# Calculate relative abundance
benthic_rel <- data_benthic %>%
  count(Benthic.Group, Benthic.Type, name = "Count") %>%
  mutate(RelAbundance = Count / sum(Count)) %>%
  arrange(Benthic.Group, desc(RelAbundance)) %>%
  mutate(
    Benthic.Type = factor(Benthic.Type, levels = unique(Benthic.Type))
  )

# Number of unique quadrats
N_quadrats <- data_benthic %>%
  distinct(Site, Transect, Quadrat) %>%
  nrow()

# Base colors by ecological group
group_bases <- c(
  Hard_Coral = "coral4",
  Algae      = "springgreen4",
  Misc       = "wheat4",
  CTB      = "black"
)

# Generate shades within each ecological group
benthic_cols <- benthic_rel %>%
  distinct(Benthic.Group, Benthic.Type) %>%
  group_by(Benthic.Group) %>%
  mutate(
    Color = lighten(
      group_bases[Benthic.Group[1]],
      amount = seq(0, 0.65, length.out = n())
    )
  ) %>%
  ungroup()

color_mapping <- setNames(
  benthic_cols$Color,
  benthic_cols$Benthic.Type
)

# Plot
benthic_overall_plot <- ggplot(
  benthic_rel,
  aes(
    x = "Overall",
    y = RelAbundance,
    fill = Benthic.Type
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(
      x = "Overall",
      y = 1.03,
      label = paste0(N_quadrats)
    ),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 4
  ) +
  scale_fill_manual(values = color_mapping) +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
)+
  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Benthic Type"
  ) +
  theme_bw()+
  theme(legend.position = "none")
benthic_overall_plot
```

Visualize Benthic Cover by Site
```{r}

# Calculate relative abundance by Shore and Site
benthic_rel_site <- data_benthic %>%
  count(Shore, Site, Benthic.Group, Benthic.Type, name = "Count") %>%
  group_by(Shore, Site) %>%
  mutate(RelAbundance = Count / sum(Count)) %>%
  ungroup() %>%
  arrange(Benthic.Group, desc(RelAbundance)) %>%
  mutate(
    Site = factor(Site, levels = site_order),
    Benthic.Type = factor(Benthic.Type, levels = unique(Benthic.Type))
  )

# Number of unique quadrats per Shore and Site
site_n <- data_benthic %>%
  distinct(Shore, Site, Transect, Quadrat) %>%
  count(Shore, Site, name = "N_quadrats") %>%
  mutate(
    Site = factor(Site, levels = site_order)
  )

benthic_site_plot <- ggplot(
  benthic_rel_site,
  aes(
    x = Site,
    y = RelAbundance,
    fill = Benthic.Type
  )
) +
  geom_col(width = 0.7) +
  geom_text(
    data = site_n,
    aes(
      x = Site,
      y = 1.03,
      label = N_quadrats
    ),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 2
  ) +
  facet_grid(~ Shore, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = color_mapping) +
  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
)+
  labs(
    x = "Site",
    y = "Relative Abundance",
  fill = NULL
  ) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "gray", color = "black"),
    strip.text = element_text(color = "black", face = "bold"),
    legend.text.align = 0,
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 6),
    legend.key.height = unit(0.3, "cm")
  )
benthic_site_plot
```

Coral only relative abundance overall
```{r}

data_benthic_coral <- data_benthic %>%
  filter(Benthic.Group=="Hard_Coral")

# Calculate relative abundance
benthic_coral_rel <- data_benthic_coral %>%
  count(Benthic.Group, Benthic.Type, name = "Count") %>%
  mutate(RelAbundance = Count / sum(Count)) %>%
  arrange(Benthic.Group, desc(RelAbundance)) %>%
  mutate(
    Benthic.Type = factor(Benthic.Type, levels = unique(Benthic.Type))
  )

# Number of unique quadrats
N_quadrats <- data_benthic_coral %>%
  distinct(Site, Transect, Quadrat) %>%
  nrow()

unique(data_benthic_coral$Benthic.Type)

coral_types <- sort(unique(data_benthic_coral$Benthic.Type))

coral_color_mapping <- setNames(
  qualitative_hcl(
    n = length(coral_types),
    palette = "Dark 3"
  ),
  coral_types
)

coral_color_mapping

# Plot
hard_coral_overall_plot <- ggplot(
  benthic_coral_rel,
  aes(
    x = "Overall",
    y = RelAbundance,
    fill = Benthic.Type
  )
) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = coral_color_mapping) +
  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
)+
  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Hard Coral Type"
  ) +
  theme_bw()+
  theme(legend.position = "none")
hard_coral_overall_plot
```
Coral only relative abundance by Site
```{r}
# Coral relative abundance by Site
benthic_coral_rel_site <- data_benthic_coral %>%
  count(Shore, Site, Benthic.Type, name = "Count") %>%
  group_by(Shore, Site) %>%
  mutate(RelAbundance = Count / sum(Count)) %>%
  ungroup() %>%
  arrange(desc(RelAbundance)) %>%
  mutate(
    Site = factor(Site, levels = site_order),
    Benthic.Type = factor(Benthic.Type, levels = coral_types)
  )

# Number of unique quadrats per Site
site_n_coral <- data_benthic_coral %>%
  distinct(Shore, Site, Transect, Quadrat) %>%
  count(Shore, Site, name = "N_quadrats") %>%
  mutate(Site = factor(Site, levels = site_order))

# Plot
hard_coral_site_plot <- ggplot(
  benthic_coral_rel_site,
  aes(
    x = Site,
    y = RelAbundance,
    fill = Benthic.Type
  )
) +
  geom_col(width = 0.7) +
  facet_grid(~ Shore, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = coral_color_mapping) +
  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
)+
  labs(
    x = "Site",
    y = "Relative Abundance",
  fill = NULL) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "gray", color = "black"),
    strip.text = element_text(color = "black", face = "bold"),
    legend.text.align = 0,
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 6),
    legend.key.height = unit(0.3, "cm")
  )
hard_coral_site_plot
```

Visualize the size and species data
```{r}

#Set the Site Order
data$Site <- factor(
  data$Site,
  levels = c(
    "S2","S3","S4","S5","S6","S9","S11","S12","S14","S15","S18"
  )
)

#Set the Species Order
data$Species <- factor(
  data$Species,
  levels = c( "P. acuta", "P. effusa","P. grandis","P. meandrina","P. tuahiniensis", "P. verrucosa"
  )
)

#Calculate relative abundance by species and site
rel_abund <- data %>%
  count(Shore, Site, Species) %>%
  group_by(Shore, Site) %>%
  mutate(RelAbundance = n / sum(n)) %>%
  ungroup()

#Count sample size
site_n <- data %>%
  count(Shore, Site, name = "N")

pocillopora_site_plot <- ggplot(rel_abund,
       aes(x = Site,
           y = RelAbundance,
           fill = Species)) +
  geom_col() +

  geom_text(
    data = site_n,
    aes(x = Site, y = 1.03, label = N),
    inherit.aes = FALSE,
    size = 2
  ) +
  labs(
  x = "Site",
  y = "Relative Abundance",
  fill = NULL)+

  facet_grid(~ Shore, scales = "free_x", space = "free_x") +

  scale_fill_manual(
    values = c(
      "P. tuahiniensis" = "#D55E00",
      "P. meandrina"    = "#0072B2",
      "P. verrucosa"    = "#E69F00",
      "P. grandis"      = "#56B4E9",
      "P. effusa"       = "#009E73",
      "P. acuta"        = "#e63946"
    ),
    labels = c(
      "P. tuahiniensis" = expression(italic("P. tuahiniensis")),
      "P. meandrina"    = expression(italic("P. meandrina")),
      "P. verrucosa"    = expression(italic("P. verrucosa")),
      "P. grandis"      = expression(italic("P. grandis")),
      "P. effusa"       = expression(italic("P. effusa")),
      "P. acuta"        = expression(italic("P. acuta"))
    )
  ) +

  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
)+
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "gray", color = "black"),
    strip.text = element_text(color = "black", face = "bold"),
    legend.text.align = 0,
    legend.position = "right",
    legend.title = element_blank(),
    legend.text = element_text(size = 6),
    legend.key.height = unit(0.4, "cm")
  )
pocillopora_site_plot

```

PERMANOVA test of relative abundance by Site
```{r}
comm <- data %>%
  count(Shore, Site, Transect, Species) %>%
  group_by(Shore, Site, Transect) %>%
  mutate(RelAbundance = n / sum(n)) %>%
  ungroup() %>%
  select(Shore, Site, Transect, Species, RelAbundance) %>%
  pivot_wider(
    names_from = Species,
    values_from = RelAbundance,
    values_fill = 0
  )

meta <- comm %>%
  select(Shore, Site, Transect)

comm_mat <- comm %>%
  select(-Shore, -Site, -Transect)

#
dist_bc <- vegdist(comm_mat, method = "bray")

site.per <- adonis2(
  dist_bc ~ Site,
  data = meta,
  permutations = 9999
)
site.per

disp <- betadisper(dist_bc, meta$Site)
anova(disp)
permutest(disp)

nmds <- metaMDS(
  comm_mat,
  distance = "bray",
  k = 2,
  trymax = 100
)

scores_df <- as.data.frame(scores(nmds, display = "sites"))

scores_df$Site <- meta$Site
scores_df$Transect <- meta$Transect
scores_df$Shore <- meta$Shore

centroids <- scores_df %>%
  group_by(Shore, Site) %>%
  summarise(
    NMDS1 = mean(NMDS1),
    NMDS2 = mean(NMDS2),
    .groups = "drop"
  )

# site_cols <- c(
#   # North - blues
#   "S2"  = "#08306B",
#   "S3"  = "#2171B5",
#   "S4"  = "#6BAED6",
#   "S5"  = "#C6DBEF",
# 
#   # East - greens
#   "S6"  = "#00441B",
#   "S9"  = "#238B45",
#   "S11" = "#C7E9C0",
# 
#   # West - oranges/reds
#   "S12"  = "#7F2704",
#   "S14" = "#D94801",
#   "S15" = "#FD8D3C",
#   "S18" = "#FDD0A2"
# )

site_cols <- c(
  # North - blues
  "S2"  = "turquoise2",
  "S3"  = "turquoise2",
  "S4"  = "turquoise2",
  "S5"  = "turquoise2",

  # East - greens
  "S6"  = "palegreen1",
  "S9"  = "palegreen1",
  "S11" = "palegreen1",

  # West - oranges/reds
  "S12"  = "plum2",
  "S14" = "plum2",
  "S15" = "plum2",
  "S18" = "plum2"
)




#plot
ggplot(scores_df,
       aes(NMDS1, NMDS2,
           color = Site,
           shape = Site)) +

  geom_polygon(
    data = scores_df %>%
      group_by(Site) %>%
      slice(chull(NMDS1, NMDS2)),
    aes(fill = Site),
    alpha = 0.3,
    color = NA
  ) +

  geom_point(
    size = 2,
    alpha = 0.8
  ) +

geom_text(
  data = centroids,
  aes(label = Site),
  color = "black",
  fontface = "bold",
  size = 3,
  nudge_y = 0.02
)+

  scale_color_manual(values = site_cols) +
  scale_fill_manual(values = site_cols) +

  scale_shape_manual(
    values = c(
      S2 = 1,  # solid circle
      S3  = 2,  # solid triangle
      S4  = 3,   # solid square
      S5 = 4,  # solid circle
      S6  = 5,  # solid triangle
      S9  = 6,   # solid square
      S11 = 7,  # solid circle
      S12  = 8,  # solid triangle
      S14  = 9,   # solid square
      S15  = 10,   # solid square
      S18  = 11   # solid square
    )
  ) +

  theme_bw()
```


Plot total Pocillopora with all sites together
```{r}
# Calculate overall relative abundance
rel_abund_overall <- data %>%
  count(Species) %>%
  mutate(RelAbundance = n / sum(n))

# Total sample size
N_total <- nrow(data)

pocillopora_overall_plot <- ggplot(rel_abund_overall,
       aes(x = "Overall",
           y = RelAbundance,
           fill = Species)) +
  geom_col() +

  geom_text(
    aes(x = "Overall",
        y = 1.03,
        label = N_total),
    inherit.aes = FALSE,
    fontface = "plain",
    size = 3.5
  ) +

  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Species"
  ) +

  scale_fill_manual(
    values = c(
      "P. tuahiniensis" = "#D55E00",
      "P. meandrina"    = "#0072B2",
      "P. verrucosa"    = "#E69F00",
      "P. grandis"      = "#56B4E9",
      "P. effusa"       = "#009E73",
      "P. acuta"        = "#e63946"
    ),
    labels = c(
      "P. tuahiniensis" = expression(italic("P. tuahiniensis")),
      "P. meandrina"    = expression(italic("P. meandrina")),
      "P. verrucosa"    = expression(italic("P. verrucosa")),
      "P. grandis"      = expression(italic("P. grandis")),
      "P. effusa"       = expression(italic("P. effusa")),
      "P. acuta"        = expression(italic("P. acuta"))
    )
  ) +

  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1)
)+
  theme_bw()+
  theme(legend.position = "none")
pocillopora_overall_plot

```

Plot multipanel figures
```{r}
# plot objects
Fig1 <- map_plot + labs(tag = "A")
Fig2 <- benthic_overall_plot + labs(tag = "B")
Fig3 <- hard_coral_overall_plot + labs(tag = "D")
Fig4 <- pocillopora_overall_plot + labs(tag = "F")

Fig5 <- benthic_site_plot + labs(tag = "C")
Fig6 <- hard_coral_site_plot + labs(tag = "E")
Fig7 <- pocillopora_site_plot + labs(tag = "G")



row1 <- wrap_plots(Fig1, nrow = 1)
row2 <- wrap_plots(Fig2, Fig5, nrow = 1)
row3 <- wrap_plots(Fig3, Fig6, nrow = 1)
row4 <- wrap_plots(Fig4, Fig7, nrow = 1)

multi_panel_fig <- row1 / row2 /row3 / row4 &
  theme(
    plot.tag = element_text(
      face = "bold",
      size = 18
    ),
    plot.tag.position = c(0.02, 0.98)
  )

ggsave(
  "Output/multi_panel_fig.pdf",
  multi_panel_fig,
  width = 10,
  height = 10,
  units = "in",
  device = cairo_pdf,
  bg = "white"
)

ggsave(
  "Output/multi_panel_fig.jpg",
  multi_panel_fig,
  width = 10,
  height = 10,
  units = "in",
  bg = "white"
)

```


Relative Abundance by Site and Size Class
```{r}
rel_abund <- data %>%
  count(Shore, Site, SizeClass, Species) %>%
  group_by(Shore, Site, SizeClass) %>%
  mutate(RelAbundance = n / sum(n)) %>%
  ungroup()

site_n <- data %>%
  count(Shore, Site, SizeClass, name = "N")


pocillopora_size_plot <- ggplot(rel_abund,
       aes(x = Site,
           y = RelAbundance,
           fill = Species)) +

  geom_col() +

  geom_text(
    data = site_n,
    aes(x = Site, y = 1.03, label = N),
    inherit.aes = FALSE,
    size = 2.5
  ) +

  #facet_grid(
  #  SizeClass ~ Shore,
  #  scales = "free_x",
  #  space = "free_x"
 # ) +

  scale_fill_manual(
    values = c(
      "P. tuahiniensis" = "#D55E00",
      "P. meandrina"    = "#0072B2",
      "P. verrucosa"    = "#E69F00",
      "P. grandis"      = "#56B4E9",
      "P. effusa"       = "#009E73",
      "P. acuta"        = "#e63946"
    ),
    labels = c(
      "P. tuahiniensis" = expression(italic("P. tuahiniensis")),
      "P. meandrina"    = expression(italic("P. meandrina")),
      "P. verrucosa"    = expression(italic("P. verrucosa")),
      "P. grandis"      = expression(italic("P. grandis")),
      "P. effusa"       = expression(italic("P. effusa")),
      "P. acuta"        = expression(italic("P. acuta"))
    )
  ) +

  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(
  breaks = seq(0, 1, by = 0.25),
  labels = scales::percent_format(accuracy = 1))+

  labs(
    x = "Site",
    y = "Relative Abundance",
    fill = NULL
  ) +

  theme_bw() +
  theme(
    strip.background = element_rect(fill = "gray", color = "black"),
    strip.text = element_text(color = "black", face = "bold"),
    legend.text.align = 0,
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

pocillopora_size_plot.2 <- pocillopora_size_plot +
  facet_grid2(
    SizeClass ~ Shore,
    scales = "free_x",
    space = "free_x",
    strip = strip_themed(
      background_x = elem_list_rect(
        fill = c(
          "turquoise2",  # North
          "palegreen1",  # East
          "plum2"        # West
        ),
        color = "black"
      ),
      text_x = elem_list_text(
        face = "bold",
        color = "black"
      )
    )
  )
pocillopora_size_plot.2
```

```{r}
# Function that runs PERMANOVA, dispersion test, NMDS, and generates a plot
run_sizeclass_nmds <- function(size_class_name) {

  data_sub <- data %>%
    filter(SizeClass == size_class_name)

  comm <- data_sub %>%
    count(Shore, Site, Transect, Species) %>%
    group_by(Shore, Site, Transect) %>%
    mutate(RelAbundance = n / sum(n)) %>%
    ungroup() %>%
    select(Shore, Site, Transect, Species, RelAbundance) %>%
    pivot_wider(
      names_from = Species,
      values_from = RelAbundance,
      values_fill = 0
    )

  meta <- comm %>%
    select(Shore, Site, Transect)

  comm_mat <- comm %>%
    select(-Shore, -Site, -Transect)

  # Bray-Curtis distance
  dist_bc <- vegdist(comm_mat, method = "bray")

  # PERMANOVA
  site_per <- adonis2(
    dist_bc ~ Site,
    data = meta,
    permutations = 9999
  )

  # Dispersion test
  disp <- betadisper(dist_bc, meta$Site)
  disp_anova <- anova(disp)
  disp_perm <- permutest(disp, permutations = 9999)

  # NMDS
  nmds <- metaMDS(
    comm_mat,
    distance = "bray",
    k = 2,
    trymax = 100
  )

  # Extract PERMANOVA statistics
  per_tab <- as.data.frame(site_per)

  site_p <- per_tab["Site", "Pr(>F)"]
  site_R2 <- per_tab["Site", "R2"]

  stat_label <- paste0(
    "p = ", signif(site_p, 3),
    #"\nR² = ", round(site_R2, 2),
    "\nStress = ", round(nmds$stress, 3)
  )

  # NMDS scores
  scores_df <- as.data.frame(scores(nmds, display = "sites"))

  scores_df$Shore <- meta$Shore
  scores_df$Site <- meta$Site
  scores_df$Transect <- meta$Transect
  scores_df$SizeClass <- size_class_name

  # Site centroids
  centroids <- scores_df %>%
    group_by(Shore, Site) %>%
    summarise(
      NMDS1 = mean(NMDS1),
      NMDS2 = mean(NMDS2),
      .groups = "drop"
    )

  # Convex hulls
  hulls <- scores_df %>%
    group_by(Site) %>%
    filter(n() >= 3) %>%
    slice(chull(NMDS1, NMDS2))

  # Plot
  p <- ggplot(
    scores_df,
    aes(
      NMDS1,
      NMDS2,
      color = Site,
      shape = Site
    )
  ) +

    geom_polygon(
      data = hulls,
      aes(fill = Site),
      alpha = 0.3,
      color = NA
    ) +

    geom_point(
      size = 2,
      alpha = 0.8
    ) +

    geom_text(
      data = centroids,
      aes(label = Site),
      color = "black",
      fontface = "bold",
      size = 2
    ) +

    annotate(
      "text",
      x = Inf,
      y = Inf,
      label = stat_label,
      hjust = 1.05,
      vjust = 1.1,
      size = 3
    ) +

    scale_color_manual(values = site_cols) +

    scale_fill_manual(values = site_cols) +

    scale_shape_manual(
      values = c(
        S2  = 1,
        S3  = 2,
        S4  = 3,
        S5  = 4,
        S6  = 5,
        S9  = 6,
        S11 = 7,
        S12 = 8,
        S14 = 9,
        S15 = 10,
        S18 = 11
      )
    ) +

    labs(
      x = "NMDS1",
      y = "NMDS2",
      color = "Site",
      fill = "Site",
      shape = "Site"
    ) +

    theme_bw() +
    theme(
      legend.position = "none"
    )

  list(
    size_class = size_class_name,
    permanova = site_per,
    dispersion_anova = disp_anova,
    dispersion_permutest = disp_perm,
    nmds = nmds,
    scores = scores_df,
    centroids = centroids,
    plot = p
  )
}

# Run for all size classes
size_results <- map(levels(data$SizeClass), run_sizeclass_nmds)
names(size_results) <- levels(data$SizeClass)

# Summary table
size_summary <- map_dfr(size_results, function(x) {

  per <- as.data.frame(x$permanova)

  tibble(
    SizeClass = x$size_class,
    Df        = per["Site", "Df"],
   # R2        = per["Site", "R2"],
    F         = per["Site", "F"],
    p.value   = per["Site", "Pr(>F)"],
    Stress    = x$nmds$stress
  )
})

size_summary

# Extract plots
pocillopora_NMDS_02 <- size_results[["0-2cm"]]$plot +
  labs(x = NULL) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

pocillopora_NMDS_25 <- size_results[[">2cm-5cm"]]$plot +
  labs(x = NULL) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

pocillopora_NMDS_514 <- size_results[[">5cm-14cm"]]$plot +
  labs(x = NULL) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none"
  )

pocillopora_NMDS_14up <- size_results[[">14cm"]]$plot +
  labs(x = "NMDS1") +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

Plot multipanel figures
```{r}
# plot objects
Fig8 <- pocillopora_size_plot.2 + labs(tag = "A")
Fig9 <- pocillopora_NMDS_02 + labs(tag = "B")
Fig10 <- pocillopora_NMDS_25 + labs(tag = "C")
Fig11 <- pocillopora_NMDS_514 + labs(tag = "D")
Fig12 <- pocillopora_NMDS_14up + labs(tag = "E")

right_column <-
  (Fig9 /
   Fig10  /
   Fig11 /
   Fig12) +
  plot_layout(guides = "collect") &
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )


multi_panel_nmds <-
  Fig8 | right_column +
  plot_layout(
    widths = c(1.5, 1),
    heights = c(1, 1, 1, 1)
  )

ggsave(
  "Output/multi_panel_nmds.pdf",
  multi_panel_nmds,
  width = 10,
  height = 14,
  units = "in",
  device = cairo_pdf,
  bg = "white"
)

ggsave(
  "Output/multi_panel_nmds.jpg",
  multi_panel_nmds,
  width = 10,
  height = 14,
  units = "in",
  bg = "white"
)

```
