---
title: "Relative_abundance"
output: html_document
---
This script aims to analyse the relative abundance of Pocillopora around the island of Mo’orea under 12 environmental conditions. This script also aims to characterize the benthos at each site in order to better understand the biodiversity in which the Pocillopora community lives.

Q1: What does the benthic cover of the lagoon around the island look like?
Q2: How are Pocillopora species distributed around the lagoon?

What corals are found on the lagoon reefs of Mo’orea?
What Pocillopora species are found on the lagoon reefs of Mo’orea?

H0: Benthic cover, coral cover, and Pocillopora cover differ by site.

#Install and load packages
```{r}
library("ggplot2")
library("vegan")
#library("ggpubr")
library("tidyverse")
```

Load Data Files for Analysis
```{r}
#read in species id data
data_haplotype <- read.csv("Data/Haplotype_data_modified_duplication.csv")

#read in colony diameter data
data_size <- read.csv("Data/POC_RA_Size.csv")
```

Data Cleaning
```{r}
#filter to only data from the first 3 transects
data_size <- data_size %>%
  filter(Transect %in% c("T1", "T2", "T3"))

# Remove uninformative columns ("Sanger.Plate.Name","Sequence.ID","Plate.Well", "gDNA.Extraction.Plate") and filter out NA rows
data_species <- data_haplotype %>%
  select(-c("Sanger.Plate.Name","Sequence.ID","Plate.Well", "gDNA.Extraction.Plate", "Notes")) %>% na.omit()

# Remove uninformative columns ("Diameter.1","Diameter.2","Diameter.3","Note") and filter out NA rows and DIV/0! rows and check if diameter is numeric
data_diameters <- data_size %>%
  select(-c("Diameter.1","Diameter.2","Diameter.3","Note")) %>%
  na.omit() %>%
  filter(Diameter.mean!="#DIV/0!") 

str(data_diameters)
data_diameters$Diameter.mean <- as.numeric(data_diameters$Diameter.mean)
str(data_diameters)

```

Quality Control of Sample.IDs
```{r}
# Check species and diameter datasets by Sample.ID to see if there are any ID problems

# Check to see they are all distinct
dup_species <- data_species %>% count(Sample.ID, sort = TRUE) %>% filter(n > 1)
dup_species$Sample.ID <- as.numeric(dup_species$Sample.ID)
  
#generate a list of duplicated IDs
dup_rows_species <- data_species %>%
  filter(Sample.ID %in% dup_species$Sample.ID) %>%
  arrange(Sample.ID)

#save a file of problematic Sample.ID to QC
write.csv(dup_rows_species, "Output/dup_rows_species.csv")

# Check to see they are all distinct
dup_diameters <-data_diameters %>% count(Sample.ID, sort = TRUE) %>% filter(n > 1)
dup_diameters$Sample.ID <- as.numeric(dup_diameters$Sample.ID)

#generate a list of duplicated IDs
dup_rows_size <- data_diameters %>%
  filter(Sample.ID %in% dup_diameters$Sample.ID) %>%
  arrange(Sample.ID)

#save a file of problematic Sample.ID to QC
write.csv(dup_rows_size, "Output/dup_rows_size.csv")

# Check to see if all Sample.ID are all numeric
bad_ids_species <- data_species$Sample.ID[
  is.na(suppressWarnings(
    as.numeric(as.character(data_species$Sample.ID))
  ))
]
bad_ids_species

#An issues come from 2 samples because _BIS is present at the end of the Sample.ID
#For these samples "_BIS" means #.2 in French, essentially a duplicate name
#We do not have pictures of these two _BIS corals so they are going to be excluded from the dataset

data_species <- data_species %>%
  filter(!grepl("^[0-9]+_BIS$", Sample.ID)) %>% #filter out _BIS samples
  mutate(Sample.ID = as.numeric(as.character(Sample.ID))) # change the sample.ID to numeric
  
# Check to see if all Sample.ID are all numeric
bad_ids_species <- data_species$Sample.ID[
  is.na(suppressWarnings(
    as.numeric(as.character(data_species$Sample.ID))
  ))
]
bad_ids_species
#There are only numeric IDs in species data

# Check to see if all Sample.ID are all numeric
bad_ids_diameters <- data_diameters$Sample.ID[
  is.na(suppressWarnings(
    as.numeric(as.character(data_diameters$Sample.ID))
  ))
]
bad_ids_diameters 
#There are only numeric IDs in diameter data

# Remove remaining duplicates if QC and manual check of the photos and notebooks do not reconcile and you cannot confirm the sample ID
data_species <- data_species %>%
  filter(!Sample.ID %in% dup_species$Sample.ID)

data_diameters <- data_diameters %>% 
  filter(!Sample.ID %in% dup_diameters$Sample.ID)

```

Sanity checks for data quality
```{r}
# This section verifies the integrity of the datasets before
# any analysis. Each check is documented for reproducibility
# and transparency for reviewers.

# --- QC 1: Dataset Dimensions -------------------------------
# Verify the number of rows and columns after cleaning
cat("data_species dimensions:  ", dim(data_species),  "\n")
cat("data_diameters dimensions:", dim(data_diameters), "\n")
#when all species IDs are complete we would expect the same number of rows in both dataframes
#We have ~300 more species than we do sizes, which is likely due to the species coming from transects 4,5,6 and we are not able to filter them by transect becasue the file did not have metadata within it


# --- QC 2: Dataset Structure --------------------------------
# Verify column names and types are as expected
#Species should be character
#Sample.ID should be numeric
#Diameter.mean should be numeric
str(data_species)
str(data_diameters)


# --- QC 3: Missing Values -----------------------------------
# Confirm no NAs remain after na.omit in cleaning step
#results from this should be all 0's
cat("\nMissing values in data_species:\n")
colSums(is.na(data_species))

cat("\nMissing values in data_diameters:\n")
colSums(is.na(data_diameters))


# --- QC 4: Duplicate Sample.IDs -----------------------------
# Confirm no duplicates remain after deduplication step
cat("\nRemaining duplicates in data_species:\n")
data_species  %>% count(Sample.ID) %>% filter(n > 1)

cat("\nRemaining duplicates in data_diameters:\n")
data_diameters %>% count(Sample.ID) %>% filter(n > 1)

# --- QC 5: Value Range & Reality Check ------------------------
# Check for impossible or unexpected values in key variables
#this should be numeric, not character
cat("\nSummary of data_diameters:\n")
range(data_diameters$Diameter.mean)
#this should range from >0-100, where 100 indicates everything >14cm

cat("\nUnique sites found in data_diameters:\n")
unique(data_diameters$Site)

cat("\nUnique transects found in data_diameters:\n")
unique(data_diameters$Transect)

cat("\nUnique species found in data_species:\n")
#should be "P. grandis" "P. meandrina" "P. verrucosa" "P. effusa" "P. tuahiniensis" "P. acuta"  
unique(data_species$Species)

cat("\nUnique haplotypes found in data_species:\n")
#should be "Haplotype 1a" "Haplotype 3"  "Haplotype 2"  "Haplotype 11" "Haplotype 10" "Haplotype 8a" "Haplotype 5a"
unique(data_species$Haplotype)


```

Join the species and size datasets together for Size Class specific analyses
```{r}
# --- QC 6: Join Integrity -----------------------------------
# Identify Sample.IDs present in one dataset but not the other
# These samples will be excluded from the merged dataset
lost_from_species <- anti_join(data_species,
  data_diameters,
  by = "Sample.ID"
)

lost_from_diameters <- anti_join(
  data_diameters,
  data_species,
  by = "Sample.ID"
)

# Store excluded Sample.IDs for traceability
excluded_from_species   <- lost_from_species  %>% pull(Sample.ID)
excluded_from_diameters <- lost_from_diameters %>% pull(Sample.ID)

cat("\nSamples excluded from data_species (no size match):",   length(excluded_from_species),   "\n")
cat("Samples excluded from data_diameters (no species match):", length(excluded_from_diameters), "\n")

# --- Remove unmatched samples from each dataset -------------
data_species <- data_species %>%
  filter(!Sample.ID %in% excluded_from_species)
dim(data_species)

data_diameters <- data_diameters %>%
  filter(!Sample.ID %in% excluded_from_diameters)
dim(data_diameters)

# --- Rebuild merged dataset after exclusions ----------------
data <- data_species %>%
  inner_join(data_diameters, by = "Sample.ID")

cat("\nFinal merged dataset dimensions after exclusions:", dim(data), "\n")

# --- Confirm no coercion warnings remain --------------------
cat("\nFinal merged dataset dimensions after exclusions:", dim(data), "\n")
cat("Missing values in final merged dataset:\n")
colSums(is.na(data))

# Check all unique species names
sort(unique(data$Species))
sort(unique(data$Haplotype))

sort(unique(data$Site))
sort(unique(data$Transect))

# Cross-check: how many transects per site?
data  %>%
  group_by(Site)  %>%
  summarise(n_transects = n_distinct(Transect),
            n_samples   = n())


# --- Assign colonies to Size Class ---------------
# Colonies recorded as 100cm represent colonies whose diameter
# could not be measured in the field (> 100cm). These are
# reclassified as a distinct ">100cm" size class rather than
# being removed, as they represent real biological observations.

data <- data %>%
  mutate(
    SizeClass = case_when(
      Diameter.mean <= 2 ~ "0-2cm",
      Diameter.mean > 2 & Diameter.mean <= 5 ~ ">2cm-5cm",
      Diameter.mean > 5 & Diameter.mean <= 14 ~ ">5cm-14cm",
      Diameter.mean > 14 ~ ">14cm",
      TRUE ~ NA_character_
    )
  ) %>%
  mutate(
    SizeClass = factor(
      SizeClass,
      levels = c("0-2cm", ">2cm-5cm", ">5cm-14cm", ">14cm")
    )
  )

#because some NA have been find into the dataset, we will remove them to avoid any trouble during the stats analysis 
# Remove rows with NA in SizeClass
data <- data %>%
  filter(!is.na(SizeClass))

# Confirm no NAs remain
cat("Remaining NAs in SizeClass:", sum(is.na(data$SizeClass)), "\n")
cat("Remaining NAs in Site:", sum(is.na(data$Site)), "\n")
cat("Remaining NAs in Transect:", sum(is.na(data$Transect)), "\n")
cat("Final dataset dimensions:", dim(data), "\n")
```


Add shore category based on site information
```{r}
#for the rest of the analysis we add the shore for species distribution analysis
# Manually assign each site to its corresponding shore
# based on the geographic position around Mo'orea island
# North: sites S1, S2, S3, S4, S5
# East:  sites S6, S9, S11
# West:  sites S12, S14, S15, S18

data <- data %>%
  mutate(Shore = case_when(
    Site %in% c("S1", "S2", "S3", "S4", "S5")       ~ "North",
    Site %in% c("S6", "S9", "S11")         ~ "East",
    Site %in% c("S12", "S14", "S15", "S18")   ~ "West",
    .default = NA_character_
  )) %>%
  mutate(Shore = factor(Shore, levels = c("North", "East", "West")))

# Verify shore assignment
data %>%
  group_by(Shore, Site) %>%
  summarise(n_samples = n(), .groups = "drop") %>%
  arrange(Shore, Site)

# Confirm no unassigned sites
cat("Samples with no shore assigned:", sum(is.na(data$Shore)), "\n")

```

Visualize the data
```{r}

#Set the Site Order
data$Site <- factor(
  data$Site,
  levels = c(
    "S2","S3","S4","S5","S6","S9","S11","S12","S14","S15","S18"
  )
)

#Set the Species Order
data$Species <- factor(
  data$Species,
  levels = c( "P. acuta", "P. effusa","P. grandis","P. meandrina","P. tuahiniensis", "P. verrucosa"
  )
)

#Calculate relative abundance by species and site
rel_abund <- data %>%
  count(Shore, Site, Species) %>%
  group_by(Shore, Site) %>%
  mutate(RelAbundance = n / sum(n)) %>%
  ungroup()

#Count sample size
site_n <- data %>%
  count(Shore, Site, name = "N")

ggplot(rel_abund,
       aes(x = Site,
           y = RelAbundance,
           fill = Species)) +
  geom_col() +

  geom_text(
    data = site_n,
    aes(x = Site, y = 1.03, label = N),
    inherit.aes = FALSE,
    size = 3
  ) +
  labs(
  x = "Site",
  y = "Relative Abundance",
  fill = "Species"
)+

  facet_grid(~ Shore, scales = "free_x", space = "free_x") +

  scale_fill_manual(
    values = c(
      "P. tuahiniensis" = "#D55E00",
      "P. meandrina"    = "#0072B2",
      "P. verrucosa"    = "#E69F00",
      "P. grandis"      = "#56B4E9",
      "P. effusa"       = "#009E73",
      "P. acuta"        = "#e63946"
    ),
    labels = c(
      "P. tuahiniensis" = expression(italic("P. tuahiniensis")),
      "P. meandrina"    = expression(italic("P. meandrina")),
      "P. verrucosa"    = expression(italic("P. verrucosa")),
      "P. grandis"      = expression(italic("P. grandis")),
      "P. effusa"       = expression(italic("P. effusa")),
      "P. acuta"        = expression(italic("P. acuta"))
    )
  ) +

  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_bw() +
  theme(
    strip.background = element_rect(fill = "gray", color = "black"),
    strip.text = element_text(color = "black", face = "bold"),
    legend.text.align = 0
  )


```

PERMANOVA test of relative abundance by Site
```{r}
comm <- data %>%
  count(Shore, Site, Transect, Species) %>%
  group_by(Shore, Site, Transect) %>%
  mutate(RelAbundance = n / sum(n)) %>%
  ungroup() %>%
  select(Shore, Site, Transect, Species, RelAbundance) %>%
  pivot_wider(
    names_from = Species,
    values_from = RelAbundance,
    values_fill = 0
  )

meta <- comm %>%
  select(Shore, Site, Transect)

comm_mat <- comm %>%
  select(-Shore, -Site, -Transect)

#
dist_bc <- vegdist(comm_mat, method = "bray")

site.per <- adonis2(
  dist_bc ~ Site,
  data = meta,
  permutations = 9999
)
site.per

disp <- betadisper(dist_bc, meta$Site)
anova(disp)
permutest(disp)

nmds <- metaMDS(
  comm_mat,
  distance = "bray",
  k = 2,
  trymax = 100
)

scores_df <- as.data.frame(scores(nmds, display = "sites"))

scores_df$Site <- meta$Site
scores_df$Transect <- meta$Transect
scores_df$Shore <- meta$Shore

centroids <- scores_df %>%
  group_by(Shore, Site) %>%
  summarise(
    NMDS1 = mean(NMDS1),
    NMDS2 = mean(NMDS2),
    .groups = "drop"
  )

site_cols <- c(
  # North - blues
  "S2"  = "#08306B",
  "S3"  = "#2171B5",
  "S4"  = "#6BAED6",
  "S5"  = "#C6DBEF",

  # East - greens
  "S6"  = "#00441B",
  "S9"  = "#238B45",
  "S11" = "#C7E9C0",

  # West - oranges/reds
  "S12"  = "#7F2704",
  "S14" = "#D94801",
  "S15" = "#FD8D3C",
  "S18" = "#FDD0A2"
)




#plot
ggplot(scores_df,
       aes(NMDS1, NMDS2, color = Site)) +

  geom_polygon(
    data = scores_df %>%
      group_by(Site) %>%
      slice(chull(NMDS1, NMDS2)),
    aes(fill = Site),
    alpha = 0.2,
    color = NA
  ) +

  geom_point(size = 3) +

  geom_label(
    data = centroids,
    aes(label = Site),
    color = "black",
    fill = "white",
    fontface = "bold",
    size = 4,
    label.size = 0.25
  ) +

  scale_color_manual(values = site_cols) +
  scale_fill_manual(values = site_cols) +

  theme_bw()
```


Plot total Pocillopora with all sites together
```{r}
# Calculate overall relative abundance
rel_abund_overall <- data %>%
  count(Species) %>%
  mutate(RelAbundance = n / sum(n))

# Total sample size
N_total <- nrow(data)

ggplot(rel_abund_overall,
       aes(x = "Overall",
           y = RelAbundance,
           fill = Species)) +
  geom_col() +

  geom_text(
    aes(x = "Overall",
        y = 1.03,
        label = N_total),
    inherit.aes = FALSE,
    fontface = "plain",
    size = 3.5
  ) +

  labs(
    x = NULL,
    y = "Relative Abundance",
    fill = "Species"
  ) +

  scale_fill_manual(
    values = c(
      "P. tuahiniensis" = "#D55E00",
      "P. meandrina"    = "#0072B2",
      "P. verrucosa"    = "#E69F00",
      "P. grandis"      = "#56B4E9",
      "P. effusa"       = "#009E73",
      "P. acuta"        = "#e63946"
    ),
    labels = c(
      "P. tuahiniensis" = expression(italic("P. tuahiniensis")),
      "P. meandrina"    = expression(italic("P. meandrina")),
      "P. verrucosa"    = expression(italic("P. verrucosa")),
      "P. grandis"      = expression(italic("P. grandis")),
      "P. effusa"       = expression(italic("P. effusa")),
      "P. acuta"        = expression(italic("P. acuta"))
    )
  ) +

  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_bw()

```

Relative Abundance by Site and Size Class
```{r}
rel_abund <- data %>%
  count(Shore, Site, SizeClass, Species) %>%
  group_by(Shore, Site, SizeClass) %>%
  mutate(RelAbundance = n / sum(n)) %>%
  ungroup()

site_n <- data %>%
  count(Shore, Site, SizeClass, name = "N")

ggplot(rel_abund,
       aes(x = Site,
           y = RelAbundance,
           fill = Species)) +

  geom_col() +

  geom_text(
    data = site_n,
    aes(x = Site, y = 1.03, label = N),
    inherit.aes = FALSE,
    size = 2.5
  ) +

  facet_grid(
    SizeClass ~ Shore,
    scales = "free_x",
    space = "free_x"
  ) +

  scale_fill_manual(
    values = c(
      "P. tuahiniensis" = "#D55E00",
      "P. meandrina"    = "#0072B2",
      "P. verrucosa"    = "#E69F00",
      "P. grandis"      = "#56B4E9",
      "P. effusa"       = "#009E73",
      "P. acuta"        = "#e63946"
    ),
    labels = c(
      "P. tuahiniensis" = expression(italic("P. tuahiniensis")),
      "P. meandrina"    = expression(italic("P. meandrina")),
      "P. verrucosa"    = expression(italic("P. verrucosa")),
      "P. grandis"      = expression(italic("P. grandis")),
      "P. effusa"       = expression(italic("P. effusa")),
      "P. acuta"        = expression(italic("P. acuta"))
    )
  ) +

  coord_cartesian(ylim = c(0, 1.08), clip = "off") +
  scale_y_continuous(labels = scales::percent_format()) +

  labs(
    x = "Site",
    y = "Relative Abundance",
    fill = "Species"
  ) +

  theme_bw() +
  theme(
    strip.background = element_rect(fill = "gray", color = "black"),
    strip.text = element_text(color = "black", face = "bold"),
    legend.text.align = 0,
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```